                  6th INTERNATIONAL COMPETITION FOR UNIVERSITY
                             STUDENTS IN MATHEMATICS
                                    Keszthely, 1999.

                                  Problems and solutions on the second day


1. Suppose that in a not necessarily commutative ring R the square of any element is 0. Prove that
abc + abc = 0 for any three elements a, b, c. (20 points)
Solution. From 0 = (a + b)2 = a2 + b2 + ab + ba = ab + ba, we have ab = −(ba) for arbitrary a, b, which
implies                                                                 
               abc = a(bc) = − (bc) a = − b(ca) = (ca)b = c(ab) = − (ab)c = −abc.


2. We throw a dice (which selects one of the numbers 1, 2, . . . , 6 with equal probability) n times. What is
the probability that the sum of the values is divisible by 5? (20 points)
                                                                                                   (r)
Solution 1. For all nonnegative integers n and modulo 5 residue class r, denote by pn the probability
                                                                                         (0)
that after n throwing the sum of values is congruent to r modulo n. It is obvious that p 0 = 1 and
 (1)    (2)   (3)    (4)
p0 = p0 = p0 = p0 = 0.
     Moreover, for any n > 0 we have
                                                 6
                                           (r)
                                                X   1 (r−i)
                                          pn =        pn−1 .                                      (1)
                                                i=1
                                                    6
                                                                                                            (r)
From this recursion we can compute the probabilities for small values of n and can conjecture that p n =
1     4                          (r)    1    1
5 + 5·6n if n ≡ r (mod )5 and pn = 5 − 5·6n otherwise. From (1), this conjecture can be proved by
induction.
Solution 2. Let S be the set of all sequences consisting of digits 1, . . . , 6 of length n. We create collections
of these sequences.
     Let a collection contain sequences of the form

                                                    . . . 6} XY1 . . . Yn−k−1 ,
                                                |66 {z
                                                    k

where X ∈ {1, 2, 3, 4, 5} and k and the digits Y1 , . . . , Yn−k−1 are fixed. Then each collection consists of 5
sequences, and the sums of the digits of sequences give a whole residue system mod 5.
    Except for the sequence 66 . . . 6, each sequence is the element of one collection. This means that the
number of the sequences, which have a sum of digits divisible by 5, is 51 (6n − 1) + 1 if n is divisible by 5,
otherwise 15 (6n − 1).
    Thus, the probability is 51 + 5·64 n if n is divisible by 5, otherwise it is 51 − 5·61 n .
Solution 3. For arbitrary positive integer k denote by pk the probability that the sum of values is k. Define
the generating function
                                       ∞                                                 n
                                       X                    x + x2 + x3 + x4 + x5 + x6
                             f (x) =         p k xk =                                          .
                                                                        6
                                       k=1

(The last equality can be easily proved by induction.)
                                      ∞
                                      P
    Our goal is to compute the sum       p5k . Let ε = cos 2π         2π
                                                            5 + i sin 5 be the first 5th root of unity. Then
                                        k=1

                                ∞
                                X              f (1) + f (ε) + f (ε2 ) + f (ε3 ) + f (ε4 )
                                      p5k =                                                .
                                                                   5
                                k=1

                                                                1
                                                     jn
      Obviously f (1) = 1, and f (εj ) = ε6n for j = 1, 2, 3, 4. This implies that f (ε) + f (ε2 ) + f (ε3 ) + f (ε4 )
                                                          ∞
                                                          P
is 64n if n is divisible by 5, otherwise it is −1
                                               6n . Thus,   p5k is 51 + 5·64 n if n is divisible by 5, otherwise it is
                                                                          k=1
1     1
5 − 5·6n .

                                                 n
                                                 P                                  n
                                                                                    P
3. Assume that x1 , . . . , xn ≥ −1 and                x3i = 0. Prove that                xi ≤ n3 . (20 points)
                                                 i=1                                i=1

Solution. The inequality
                                                                         2
                                                     3   1              1
                                             0 ≤ x3 − x + = (x + 1) x −
                                                     4   4              2
holds for x ≥ −1.
    Substituting x1 , . . . , xn , we obtain
                            n                                   n                n                n
                            X              3     1                X             3X        n      3X        n
                       0≤             x3i − xi +              =         x3i −         xi + = 0 −       xi + ,
                            i=1
                                           4     4                i=1
                                                                                4 i=1     4      4 i=1     4

     n
     P
so         xi ≤ n3 .
     i=1

Remark. Equailty holds only in the case when n = 9k, k of the x1 , ..., xn are −1, and 8k of them are 21 .
                                                                                                    
4. Prove that there exists no function f : (0, +∞) → (0, +∞) such that f 2 (x) ≥ f (x + y) f (x) + y for any
x, y > 0. (20 points)
Solution. Assume that such a function exists. The initial inequality can be written in the form f (x) −
                       2
                         (x)      (x)y
f (x + y) ≥ f (x) − ff(x)+y  = ff(x)+y . Obviously, f is a decreasing function. Fix x > 0 and choose n ∈ N such
that nf (x + 1) ≥ 1. For k = 0, 1, . . . , n − 1 we have
                                                                                         
                                             k                   k+1                f x + nk       1
                                  f       x+         −f       x+                ≥              ≥    .
                                             n                    n               nf x + nk + 1   2n

The additon of these inequalities gives f (x + 1) ≤ f (x) − 21 . From this it follows that f (x + 2m) ≤ f (x) − m
for all m ∈ N. Taking m ≥ f (x), we get a contradiction with the conditon f (x) > 0.

5. Let S be the set of all words consisting of the letters x, y, z, and consider an equivalence relation ∼ on S
satisfying the following conditions: for arbitrary words u, v, w ∈ S
        (i) uu ∼ u;
       (ii) if v ∼ w, then uv ∼ uw and vu ∼ wu.
Show that every word in S is equivalent to a word of length at most 8. (20 points)
Solution. First we prove the following lemma: If a word u ∈ S contains at least one of each letter, and
v ∈ S is an arbitrary word, then there exists a word w ∈ S such that uvw ∼ u.
       If v contains a single letter, say x, write u in the form u = u1 xu2 , and choose w = u2 . Then uvw =
(u1 xu2 )xu2 = u1 ((xu2 )(xu2 )) ∼ u1 (xu2 ) = u.
       In the general case, let the letters of v be a1 , . . . , ak . Then one can choose some words w1 , . . . , wk such
that (ua1 )w1 ∼ u, (ua1 a2 )w2 ∼ ua1 , . . . , (ua1 . . . ak )wk ∼ ua1 . . . ak−1 . Then u ∼ ua1 w1 ∼ ua1 a2 w2 w1 ∼
. . . ∼ ua1 . . . ak wk . . . w1 = uv(wk . . . w1 ), so w = wk . . . w1 is a good choice.
       Consider now an arbitrary word a, which contains more than 8 digits. We shall prove that there is a
shorter word which is equivalent to a. If a can be written in the form uvvw, its length can be reduced by
uvvw ∼ uvw. So we can assume that a does not have this form.
       Write a in the form a = bcd, where b and d are the first and last four letter of a, respectively. We prove
that a ∼ bd.

                                                                          2
    It is easy to check that b and d contains all the three letters x, y and z, otherwise their length could be
reduced. By the lemma there is a word e such that b(cd)e ∼ b, and there is a word f such that def ∼ d.
Then we can write
                             a = bcd ∼ bc(def ) ∼ bc(dedef ) = (bcde)(def ) ∼ bd.

Remark. Of course, it is enough to give for every word of length 9 an shortest shorter word. Assuming that
the first letter is x and the second is y, it is easy (but a little long) to check that there are 18 words of length
9 which cannot be written in the form uvvw.
     For five of these words there is a 2-step solution, for example

                               xyxzyzx zy ∼ xy xzyz xzyzy ∼ xyx zy zy ∼ xyxzy.

     In the remaining 13 cases we need more steps. The general algorithm given by the Solution works
for these cases as well, but needs also very long words. For example, to reduce the length of the word
a = xyzyxzxyz, we have set b = xyzy, c = x, d = zxyz, e = xyxzxzyxyzy, f = zyxyxzyxzxzxzxyxyzxyz.
The longest word in the algorithm was

                bcdedef = xyzyxzxyzxyxzxzyxyzyzxyzxyxzxzyxyzyzyxyxzyxzxzxzxyxyzxyz,

which is of length 46. This is not the shortest way: reducing the length of word a can be done for example
by the following steps:

xyzyxzx yz ∼ xyzyxz xyzy z ∼ xyzyxzxy zyx yzyz ∼ xyzyxz xyzyxz yx yz yz ∼ xy zyx zyx yz ∼ xyzyxyz.

(The last example is due to Nayden Kambouchev from Sofia University.)

                                                      1
6. Let A be a subset of Zn = Z/nZ containing at most 100 ln n elements. Define the rth Fourier coefficient
of A for r ∈ Zn by                                          
                                              X        2πi
                                      f (r) =    exp       sr .
                                                        n
                                                           s∈A

Prove that there exists an r 6= 0, such that f (r) ≥ |A|
                                                      2 . (20 points)

Solution. Let A = {a1 , . . . , ak }. Consider the k-tuples
                                                                  
                                   2πia1 t               2πiak t
                             exp           , . . . , exp               ∈ Ck ,         t = 0, 1, . . . , n − 1.
                                     n                     n

Each component is in the unit circle |z| = 1. Split the circle into 6 equal arcs. This induces a decomposition
                                                         1
of the k-tuples into 6k classes. By the condition k ≤ 100   ln n we have n > 6k , so there are two k-tuples in
the same class say for t1 < t2 . Set r = t2 − t1 . Then
                                                                                     
                                    2πiaj r                  2πaj t2   2πaj t1                    π   1
                             Re exp         = cos                    −                    ≥ cos     =
                                      n                        n         n                        3   2

for all j, so
                                                                                k
                                                    |f (r)| ≥ Re f (r) ≥          .
                                                                                2




                                                                   3
