     10th International Mathematical Competition for University Students
                            Cluj-Napoca, July 2003

                                               Day 2


1. Let A and B be n × n real matrices such that AB + A + B = 0. Prove that AB = BA.

Solution. Since (A + I)(B + I) = AB + A + B + I = I (I is the identity matrix), matrices
A + I and B + I are inverses of each other. Then (A + I)(B + I) = (B + I)(A + I) and
AB + BA.

2. Evaluate the limit                 Z 2x
                                             sinm t
                                lim                 dt (m, n ∈ N).
                               x→0+    x       tn

                                    sin t                                               sin t
Solution. We use the fact that             is decreasing in the interval (0, π) and lim       = 1.
                                      t                                            t→0+0 t
                                                 sin 2x      sin t
For all x ∈ (0, π2 ) and t ∈ [x, 2x] we have            x<         < 1, thus
                                                    2          t
                                 m Z 2x m Z 2x                     Z 2x m
                                                        sinm t
                        
                          sin 2x           t                              t
                                             n
                                               <           n
                                                               dt <          dt,
                            2x        x     t        x    t           x   tn
                                 Z 2x m                  Z 2
                                      t           m−n+1
                                        n
                                          dt = x              um−n du.
                                  x   t                    1
                      m
              sin 2x
The factor                 tends to 1. If m − n + 1 < 0, the limit of xm−n+1 is infinity; if
                 2x
                                                               R2
m − n + 1 > 0 then 0. If m − n + 1 = 0 then xm−n+1 1 um−n du = ln 2. Hence,
                                                    
                                  Z2x m             0,
                                                              m≥n
                                     sin t
                             lim              dt = ln 2, n − m = 1 .
                            x→0+0       tn          
                                                       +∞, n − m > 1.
                                                    
                                  x



3. Let A be a closed subset of Rn and let B be the set of all those points b ∈ Rn for which
there exists exactly one point a0 ∈ A such that

                                      |a0 − b| = inf |a − b|.
                                                  a∈A

Prove that B is dense in Rn ; that is, the closure of B is Rn .

Solution. Let b0 ∈
                 / A (otherwise b0 ∈ A ⊂ B), % = inf |a − b0 |. The intersection of the ball
                                                        a∈A
of radius % + 1 with centre b0 with set A is compact and there exists a0 ∈ A: |a0 − b0 | = %.

                                                  1
    Denote by Br (a) = {x ∈ Rn : |x − a| ≤ r} and ∂Br (a) = {x ∈ Rn : |x − a| = r} the
ball and the sphere of center a and radius r, respectively.
    If a0 is not the unique nearest point then for
                                                 T any point a on the open line segment (a0 , b0 )
we have B|a−a0 | (a) ⊂ B% (b0 ) and ∂B|a−a0 | (a) ∂B% (b0 ) = {a0 }, therefore (a0 , b0 ) ⊂ B and
b0 is an accumulation point of set B.

4. Find all positive integers n for which there exists a family F of three-element subsets
of S = {1, 2, . . . , n} satisfying the following two conditions:

  (i) for any two different elements a, b ∈ S, there exists exactly one A ∈ F containing
      both a, b;

 (ii) if a, b, c, x, y, z are elements of S such that if {a, b, x}, {a, c, y}, {b, c, z} ∈ F, then
      {x, y, z} ∈ F.


Solution. The condition (i) of the problem allows us to define a (well-defined) operation
∗ on the set S given by

                      a ∗ b = c if and only if {a, b, c} ∈ F, where a 6= b.

We note that this operation is still not defined completely (we need to define a ∗ a), but
nevertheless let us investigate its features. At first, due to (i), for a 6= b the operation
obviously satisfies the following three conditions:
          (a) a 6= a ∗ b 6= b;
          (b) a ∗ b = b ∗ a;
          (c) a ∗ (a ∗ b) = b.
    What does the condition (ii) give? It claims that
          (e’) x ∗ (a ∗ c) = x ∗ y = z = b ∗ c = (x ∗ a) ∗ c
for any three different x, a, c, i.e. that the operation is associative if the arguments are
different. Now we can complete the definition of ∗. In order to save associativity for non-
different arguments, i.e. to make b = a ∗ (a ∗ b) = (a ∗ a) ∗ b hold, we will add to S an extra
element, call it 0, and define
          (d) a ∗ a = 0 and a ∗ 0 = 0 ∗ a = a.
    Now it is easy to check that, for any a, b, c ∈ S ∪ {0}, (a),(b),(c) and (d), still hold, and
          (e) a ∗ b ∗ c := (a ∗ b) ∗ c = a ∗ (b ∗ c).
    We have thus obtained that (S ∪ {0}, ∗) has the structure of a finite Abelian group,
whose elements are all of order two. Since the order of every such group is a power of 2,
we conclude that |S ∪ {0}| = n + 1 = 2m and n = 2m − 1 for some integer m ≥ 1.
    Given n = 2m −1, according to what we have proven till now, we will construct a family
of three-element subsets of S satisfying (i) and (ii). Let us define the operation ∗ in the
following manner:
         if a = a0 + 2a1 + . . . + 2m−1 am−1 and b = b0 + 2b1 + . . . + 2m−1 bm−1 , where ai , bi
are either 0 or 1, we put a ∗ b = |a0 − b0 | + 2|a1 − b1 | + . . . + 2m−1 |am−1 − bm−1 |.

                                                2
   It is simple to check that this ∗ satisfies (a),(b),(c) and (e’). Therefore, if we include in
F all possible triples a, b, a ∗ b, the condition (i) follows from (a),(b) and (c), whereas the
condition (ii) follows from (e’)
   The answer is: n = 2m − 1.

5. (a) Show that for each function f : Q × Q → R there exists a function g : Q → R such
that f (x, y) ≤ g(x) + g(y) for all x, y ∈ Q.
    (b) Find a function f : R × R → R for which there is no function g : R → R such that
f (x, y) ≤ g(x) + g(y) for all x, y ∈ R.

Solution. a) Let ϕ : Q → N be a bijection. Define g(x) = max{|f (s, t)| : s, t ∈ Q, ϕ(s) ≤
ϕ(x), ϕ(t) ≤ ϕ(x)}. We have f (x, y) ≤ max{g(x), g(y)} ≤ g(x) + g(y).
                                                                 1
    b) We shall show that the function defined by f (x, y) = |x−y|   for x 6= y and f (x, x) = 0
satisfies the problem. If, by contradiction there exists a function g as above, it results, that
           1
g(y) ≥ |x−y|  − f (x) for x, y ∈ R, x 6= y; one obtains that for each x ∈ R, lim g(y) = ∞.
                                                                                y→x
We show, that there exists no function g having an infinite limit at each point of a bounded
and closed interval [a, b].
    For each k ∈ N+ denote Ak = {x ∈ [a, b] : |g(x)| ≤ k}.
    We have obviously [a, b] = ∪∞ k=1 Ak . The set [a, b] is uncountable, so at least one of the
sets Ak is infinite (in fact uncountable). This set Ak being infinite, there exists a sequence
in Ak having distinct terms. This sequence will contain a convergent subsequence (xn )n∈N
convergent to a point x ∈ [a, b]. But lim g(y) = ∞ implies that g(xn ) → ∞, a contradiction
                                      y→x
because |g(xn )| ≤ k, ∀n ∈ N.

Second solution for part (b). Let S be the set of all sequences of real numbers. The
                                    2
cardinality of S is |S| = |R|ℵ0 = 2ℵ0 = 2ℵ0 = |R|. Thus, there exists a bijection h : R → S.
Now define the function f in the following way. For any real x and positive integer n,
let f (x, n) be the nth element of sequence h(x). If y is not a positive integer then let
f (x, y) = 0. We prove that this function has the required property.
    Let g be an arbitrary R → R function. We show that there exist real numbers x, y
such that f (x, y) > g(x) + g(y). Consider the sequence (n + g(n))∞ n=1 . This sequence is an
                               ∞
element of S, thus (n + g(n))n=1 = h(x) for a certain real x. Then for an arbitrary positive
integer n, f (x, n) is the nth element, f (x, n) = n + g(n). Choosing n such that n > g(x),
we obtain f (x, n) = n + g(n) > g(x) + g(n).

6. Let (an )n∈N be the sequence defined by
                                                  n
                                             1 X        ak
                            a0 = 1, an+1 =                     .
                                           n + 1 k=0 n − k + 2

Find the limit                                    n
                                                  X ak
                                            lim            ,
                                         n→∞
                                                  k=0
                                                      2k

                                                  3
if it exists.

Solution. Consider the generating function f (x) = ∞              n
                                                         P
                                                          n=0 an x . By induction 0 < an ≤ 1,
thus this series is absolutely convergent for |x| < 1, f (0) = 1 and the function is positive
in the interval [0, 1). The goal is to compute f ( 12 ).
    By the recurrence formula,
                              ∞                       ∞ X
                                                        n
                     0
                              X
                                               n
                                                      X          ak
                    f (x) =         (n + 1)an+1 x =                     xn =
                              n=0                     n=0 k=0
                                                              n − k + 2

                           ∞             ∞             ∞
                           X
                                     k   xn−k
                                         X            X    xm
                         =     ak x           = f (x)         .
                           k=0      n=k
                                        n−k+2         m=0
                                                          m+2
Then                                           Z x      ∞
                                                 f0   X        xm+1
             ln f (x) = ln f (x) − ln f (0) =       =                    =
                                              0 f     m=0
                                                          (m + 1)(m + 2)
      ∞                                         ∞
           xm+1         xm+1                   1 X xm+1
                                                                        
     X                                                                   1        1
   =                −             =1+ 1−                       =1+ 1−        ln       ,
     m=0
          (m  + 1)    (m  +  2)               x   m=0
                                                       (m + 1)           x      1 − x
                                          
                                           1
                                    ln f      = 1 − ln 2,
                                           2
            1   e
and thus f ( ) = .
            2   2




                                                4
