                        IMC2011, Blagoevgrad, Bulgaria
                                       Day 2, July 31, 2011


                                           1
Problem 1. Let (an )∞
                    n=0 be a sequence with 2 < an < 1 for all n ≥ 0. Define the sequence (xn )n=0
                                                                                              ∞

by
                                              an+1 + xn
                          x0 = a0 ,   xn+1 =                (n ≥ 0).
                                              1 + an+1 xn
What are the possible values of lim xn ? Can such a sequence diverge?
                                  n→∞
                                                                                                (10 points)

Problem 2. An alien race has three genders: male, female, and emale. A married triple consists
of three persons, one from each gender, who all like each other. Any person is allowed to belong to
at most one married triple. A special feature of this race is that feelings are always mutual — if x
likes y, then y likes x.
    The race is sending an expedition to colonize a planet. The expedition has n males, n females,
and n emales. It is known that every expedition member likes at least k persons of each of the
two other genders. The problem is to create as many married triples as possible to produce healthy
offspring so the colony could grow and prosper.

  a) Show that if n is even and k = n2 , then it might be impossible to create even one married triple.

  b) Show that if k ≥ 3n4
                          , then it is always possible to create n disjoint married triples, thus marrying
     all of the expedition members.

                                                                                                (10 points)

Problem 3. Determine the value of
                     ∞                                         
                    X             1             1               1
                         ln 1 +       · ln 1 +      · ln 1 +          .
                    n=1
                                  n            2n            2n + 1

                                                                                                (10 points)

                                                                                       f (k) − f (m)
Problem 4. Let f (x) be a polynomial with real coefficients of degree n. Suppose that
                                                                                           k−m
is an integer for all integers 0 ≤ k < m ≤ n. Prove that a − b divides f (a) − f (b) for all pairs of
distinct integers a and b.
                                                                                         (10 points)

Problem 5. Let F = A0 A1 . . . An be a convex polygon in the plane. Define for all 1 ≤ k ≤ n − 1
the operation fk which replaces F with a new polygon

                                   fk (F ) = A0 . . . Ak−1 A′k Ak+1 . . . An ,

where A′k is the point symmetric to Ak with respect to the perpendicular bisector of Ak−1 Ak+1 . Prove
that (f1 ◦ f2 ◦ . . . ◦ fn−1 )n (F ) = F . We suppose that all operations are well-defined on the polygons,
to which they are applied, i.e. results are convex polygons again. (A0 , A1 , . . . , An are the vertices of
F in consecutive order.)
                                                                                                (10 points)
