                  IMC 2012, Blagoevgrad, Bulgaria
                              Day 1, July 28, 2012




Problem 1. For every positive integer n, let p(n) denote the number of ways to express
n as a sum of positive integers. For instance, p(4) = 5 because

                     4 = 3 + 1 = 2 + 2 = 2 + 1 + 1 = 1 + 1 + 1 + 1.

Also define p(0) = 1.
   Prove that p(n) − p(n − 1) is the number of ways to express n as a sum of integers
each of which is strictly greater than 1.
                                                                         (10 points)


Problem 2. Let n be a fixed positive integer. Determine the smallest possible rank of
an n × n matrix that has zeros along the main diagonal and strictly positive real numbers
off the main diagonal.
                                                                               (10 points)


Problem 3. Given an integer n > 1, let Sn be the group of permutations of the numbers
1, 2, . . . , n. Two players, A and B, play the following game. Taking turns, they select
elements (one element at a time) from the group Sn . It is forbidden to select an element
that has already been selected. The game ends when the selected elements generate the
whole group Sn . The player who made the last move loses the game. The first move is
made by A. Which player has a winning strategy?
                                                                               (10 points)


Problem 4. Let f : R → R be a continuously differentiable function that satisfies
f ′ (t) > f (f (t)) for all t ∈ R. Prove that f (f (f (t))) ≤ 0 for all t ≥ 0.
                                                                               (10 points)


Problem 5. Let a be a rational number and let n be a positive integer. Prove that the
              n        n
polynomial X 2 (X + a)2 + 1 is irreducible in the ring Q[X] of polynomials with rational
coefficients.
                                                                             (10 points)
