                  IMC 2014, Blagoevgrad, Bulgaria


                              Day 1, July 31, 2014




Problem 1.   Determine all pairs (a, b) of real numbers for whi h there exists a unique
symmetri 2 × 2 matrix M with real entries satisfying trace(M) = a and det(M) = b.
                              (Proposed by Stephan Wagner, Stellenbos h University)
Solution 1.   Let the matrix be              
                                         x z
                                      M=       .
                                         z y
The two onditions give us x + y = a and xy − z 2 = b. Sin e this is symmetri in x and
y , the matrix an only be unique if x = y . Hen e 2x = a and x2 − z 2 = b. Moreover,
if (x, y, z) solves the system of equations, so does (x, y, −z). So M an only be unique if
z = 0. This means that 2x = a and x2 = b, so a2 = 4b.
     If this is the ase, then M is indeed unique: if x + y = a and xy − z 2 = b, then

                  (x − y)2 + 4z 2 = (x + y)2 + 4z 2 − 4xy = a2 − 4b = 0,

so we must have x = y and z = 0, meaning that
                                                      
                                       a/2 0
                                    M=
                                        0 a/2

is the only solution.
Solution 2. Note that trace(M) = a and det(M) = b if and only if the two eigenvalues
λ1 and λ2 of M are solutions of x2 − ax + b = 0. If λ1 6= λ2 , then
                                                                
                            λ1 0                           λ 0
                      M1 =                    and     M2 = 2
                             0 λ2                          0 λ1

are two distin t solutions, ontradi ting uniqueness. Thus λ1 = λ2 = λ = a/2, whi h
implies a2 = 4b on e again. In this ase, we use the fa t that M has to be diagonalisable
as it is assumed to be symmetri . Thus there exists a matrix T su h that
                                                     
                                         −1      λ 0
                                  M =T        ·         · T,
                                                  0 λ

however this redu es to M = λ(T −1 · I · T ) = λI , whi h shows again that M is unique.
Problem 2.     Consider the following sequen e
                    (an )∞
                         n=1 = (1, 1, 2, 1, 2, 3, 1, 2, 3, 4, 1, 2, 3, 4, 5, 1, . . . ).
                                                                               n
                                                                               P
                                                                                       ak
Find all pairs (α, β) of positive real numbers su h that          = β.  lim k=1
                                                                       n→∞ nα
                                   (Proposed by Tomas Barta, Charles University, Prague)
         Let Nn = n+1     (then aNn is the rst appearan e of number n in the sequen e)
                             
Solution.
                       2
and onsider limit of the subsequen e
         PNn           Pn                          Pn    k+1
                                                                       n+2
                                                                                 1 3
            k=1 ak       k=1 1 + · · · + k           k=1                            n (1 + 2/n)(1 + 1/n)
bNn :=             =         n+1 α
                                           =             2
                                                      n+1 α
                                                                 =     n+1
                                                                          3
                                                                              α = 6
                                                                                                         .
            Nnα                                                                    (1/2)α n2α (1 + 1/n)α
                                                                           
                                 2                     2                 2

We an see that lim bNn is positive and nite if and only if α = 3/2. In this ase the
                  n→∞√                                         √
limit is equal to β = 32 . So, this pair (α, β) = ( 23 , 32 ) is the only andidate for solution.
We will show onvergen e of the original sequen e for these values of α and β .
   Let N be a positive integer in [Nn +1, Nn+1 ], i.e., N = Nn +m for some 1 ≤ m ≤ n+1.
Then we have                                n+2
                                                     m+1
                                                           
                                                      3
                                                         +    2
                                            bN =              3/2
                                                     n+1
                                                         
                                                       2
                                                           + m
whi h an be estimated by
                                        n+2                  n+2
                                                                  + n+1
                                                                             
                                         3                    3        2
                                  n+1
                                          3/2 ≤ bN ≤          n+1 3/2
                                                                                  .
                                   2
                                         +n                      2
                                        √
Sin e both bounds onverge to 32 , the sequen e bN has the same limit and we are done.


Problem 3.             Let n be a positive integer. Show that there are positive real numbers
a0 , a1 , . . . , an su h that for ea h hoi e of signs the polynomial

                                 ±an xn ± an−1 xn−1 ± · · · ± a1 x ± a0

has n distin t real roots.
                                                   (Proposed by Stephan Neupert, TUM, Mün hen)
Solution. We pro eed by indu tion on n. The statement is trivial for n = 1. Thus
assume that we have some an , . . . , a0 whi h satisfy the onditions for some n. Consider
now the polynomials
                         P̃ (x) = ±an xn+1 ± an−1 xn ± . . . ± a1 x2 ± a0 x

By indu tion hypothesis and a0 6= 0, ea h of these polynomials has n + 1 distin t zeros,
in luding the n nonzero roots of ±an xn ± an−1 xn−1 ± . . . ± a1 x ± a0 and 0. In parti ular
none of the polynomials has a root whi h is a lo al extremum. Hen e we an hoose some
ε > 0 su h that for ea h su h polynomial P̃ (x) and ea h of its lo al extrema s we have
|P̃ (s)| > ε. We laim that then ea h of the polynomials

                       P (x) = ±an xn+1 ± an−1 xn ± . . . ± a1 x2 ± a0 x ± ε
has exa tly n + 1 distin t zeros as well. As P̃ (x) has n + 1 distin t zeros, it admits a
lo al extremum at n points. Call these lo al extrema −∞ = s0 < s1 < s2 < . . . < sn <
sn+1 = ∞. Then for ea h i ∈ {0, 1, . . . , n} the values P̃ (si ) and P̃ (si+1 ) have opposite
signs (with the obvious onvention at innity). By hoi e of ε the same holds true for
P (si ) and P (si+1 ). Hen e there is at least one real zero of P (x) in ea h interval (si , si+1 ),
i.e. P (x) has at least (and therefore exa tly) n+ 1 zeros. This shows that we have found a
set of positive reals a′n+1 = an , a′n = an−1 , . . . , a′1 = a0 , a′0 = ε with the desired properties.


Problem 4.       Let n > 6 be a perfe t number, and let n = pe11 · · · pekk be its prime
fa torisation with 1 < p1 < . . . < pk . Prove that e1 is an even number.
    A number n is perfe t if s(n) = 2n, where s(n) is the sum of the divisors of n.
                          (Proposed by Javier Rodrigo, Universidad Ponti ia Comillas)
Solution.        Suppose that e1Qis odd, ontrary to the statement.
      We know that s(n) = ki=1 (1 + pi + p2i + · · · + pei i ) = 2n = 2pe11 · · · pekk . Sin e e1 is
an odd number, p1 + 1 divides the rst fa tor 1 + p1 + p21 + · · · + pe11 , so p1 + 1 divides
2n. Due to p1 + 1 > 2, at least one of the primes p1 , . . . , pk divides p1 + 1. The primes
p3 , . . . , pk are greater than p1 + 1 and p1 annot divide p1 + 1, so p2 must divide p1 + 1.
Sin e p1 + 1 < 2p2 , this possible only if p2 = p1 + 1, therefore p1 = 2 and p2 = 3. Hen e,
6|n.
      Now n, n2 , n3 , n6 and 1 are distin t divisors of n, so
                                        n n n
                          s(n) ≥ n +     + + + 1 = 2n + 1 > 2n,
                                        2 3 6
 ontradi tion.


Remark. It is well-known that all even perfe t numbers have the form n = 2
                                                                                     p−1 (2p − 1) su h
            p
that p and 2 − 1 are primes.
                                                                           p
                                 So if e1 is odd then k = 2, p1 = 2, p2 = 2 − 1, e1 = p − 1 and

e2 = 1. If n > 6 then p > 2 so p is odd and e1 = p − 1 should be even.


Problem 5.      Let A1 A2 . . . A3n be a losed broken line onsisting of 3n line segments
in the Eu lidean plane. Suppose that no three of its verti es are ollinear, and for
ea h index i = 1, 2, . . . , 3n, the triangle Ai Ai+1 Ai+2 has ounter lo kwise orientation and
∠Ai Ai+1 Ai+2 = 60◦ , using the notation A3n+1 = A1 and A3n+2 = A2 . Prove that the
                                                               3
number of self-interse tions of the broken line is at most n2 − 2n + 1.
                                                                     2
                                                           A3
                                                 A6




                                            A4        A5
                                       A1                       A2

                                                                         (Proposed by Martin Langer)
Solution.   Pla e the broken line inside an equilateral triangle T su h that their sides are
parallel to the segments of the broken line. For every i = 1, 2, . . . , 3n, denote by xi the
distan e between the segment Ai Ai+1 and that side of T whi h is parallel to Ai Ai+1 . We
will use indi es modulo 3n everywhere.
    It is easy to see that if i ≡ j (mod 3) then the polylines Ai Ai+1 Ai+2 and Aj Aj+1 Aj+2
interse t at most on e, and this is possible only if either xi < xi+1 and xj > xj+1 or
xi < xi+1 and xj > xj+1 . Moreover, su h ases over all self-interse tions. So, the number
of self-interse tions annot ex eed number of pairs (i, j) with the property

(∗) i ≡ j   (mod 3),        and (xi < xi+1 and xj > xj+1 ) or (xi > xi+1 and xj < xj+1 ).




                            Ai+2                                             Aj+2

                     xi+2              xi+1                             Ai+2         xi+1
                                                                                       xj+1
                Ai+3                     Ai+4                      Aj
                                                                                       Aj+1
                Ai                     Ai+1
                       xi                                        Ai             xj    Ai+1
                              xi+3                                      xi

     Grouping the indi es 1, 2, . . . , 3n, by remainders modulo 3, we have n indi es in ea h
residue lass. Altogether there are 3 n2 index pairs (i, j) with i ≡ j (mod 3). We will
show that for every integer k with 1 ≤ k < n2 , there is some index i su h that the pair
(i, i + 6k) does not satisfy (∗). This is already n−1   2
                                                            pair; this will prove that there are
at most                                        
                                        n   n−1  3
                                   3      −     ≥ n2 − 2n + 1
                                        2    2   2
self-interse tions.
     Without loss of generality we may assume that x3n = x0 is the smallest among
x1 , . . . , x3n . Suppose that all of the pairs

         (−6k, 0),     (−6k + 1, 1),            (−6k + 2, 2),   ...,    (−1, 6k − 1),         (0, 6k)   (∗∗)

satisfy (∗). Sin e x0 is minimal, we have x−6k > x0 . The pair (−6k, 0) satises (∗), so
x−6k+1 < x1 . Then we an see that x−6k+2 > x2 , and so on; nally we get x0 > x6k .
But this ontradi ts the minimality of x0 . Therefore, there is a pair in (**) that does not
satisfy (∗).

                            n   n−1   3 2        
Remark. The bound 3
                            2 −    2   = 2 n − 2n + 1 is sharp.
