                   IMC2011, Blagoevgrad, Bulgaria
                                   Day 1, July 30, 2011


Problem 1. Let f : R → R be a continuous function. A point x is called a shadow point
if there exists a point y ∈ R with y > x such that f (y) > f (x). Let a < b be real numbers
and suppose that
   • all the points of the open interval I = (a, b) are shadow points;
   • a and b are not shadow points.
Prove that
   a) f (x) ≤ f (b) for all a < x < b;
   b) f (a) = f (b).
                                                                                        (10 points)

Problem 2. Does there exist a real 3 × 3 matrix A such that tr(A) = 0 and A2 + At = I?
(tr(A) denotes the trace of A, At is the transpose of A, and I is the identity matrix.)
                                                                                (10 points)

Problem 3. Let p be a prime number. Call a positive integer n interesting if
                                xn − 1 = (xp − x + 1)f (x) + pg(x)
for some polynomials f and g with integer coefficients.
    a) Prove that the number pp − 1 is interesting.
    b) For which p is pp − 1 the minimal interesting number?
                                                                                        (10 points)

Problem 4. Let A1 , A2 , . . . , An be finite, nonempty sets. Define the function
                                X
                                n           X
                      f (t) =                         (−1)k−1 t|Ai1 ∪Ai2 ∪...∪Aik | .
                                k=1 1≤i1 <i2 <...<ik ≤n

Prove that f is nondecreasing on [0, 1].
   (|A| denotes the number of elements in A.)
                                                                                        (10 points)

Problem 5. Let n be a positive integer and let V be a (2n − 1)-dimensional vector space
over the two-element field. Prove that for arbitrary vectors v1 , . . . , v4n−1 ∈ V , there exists
a sequence 1 ≤ i1 < . . . < i2n ≤ 4n − 1 of indices such that vi1 + . . . + vi2n = 0.
                                                                                       (10 points)
