              Solutions for problems in the
      9th International Mathematics Competition
                for University Students
                    Warsaw, July 19 - July 25, 2002

                                 Second Day


Problem 1. Compute the determinant of the n × n matrix A = [a ij ],
                        (
                          (−1)|i−j| , if i 6= j,
                  aij =
                          2,          if i = j.

Solution. Adding the second row to the first one, then adding the third row
to the second one, ..., adding the nth row to the (n − 1)th, the determinant
does not change and we have


         2 −1 +1 . . .          ±1 ∓1    1  1   0  0               ...    0    0
         −1 2 −1 . . .          ∓1 ±1    0  1   1  0               ...    0    0
         +1 −1 2 . . .          ±1 ∓1    0  0   1  1               ...    0    0
det(A) = .   .. .. ..            .. .. = ..  .. .. ..              ..     ..   .. .
          ..  .  .    .           .  .    .   .  .  .                 .    .    .
         ∓1 ±1 ∓1 . . .          2 −1    0  0   0  0               ...     1 1
         ±1 ∓1 ±1 . . .         −1 2     ±1 ∓1 ±1 ∓1               ...    −1 2
    Now subtract the first column from the second, then subtract the result-
ing second column from the third, ..., and at last, subtract the (n − 1)th
column from the nth column. This way we have

                         1 0 0         ...    0    0
                         0 1 0         ...    0    0
                det(A) = ... ... ...   ..
                                          .
                                              ..
                                               .
                                                   ..
                                                    .   = n + 1.
                         0 0 0         ...    1  0
                         0 0 0         ...    0 n+1

Problem 2. Two hundred students participated in a mathematical con-
test. They had 6 problems to solve. It is known that each problem was
correctly solved by at least 120 participants. Prove that there must be two
participants such that every problem was solved by at least one of these two
students.
Solution. For each pair of students, consider
                                             the set of those problems which
was not solved by them. There exist 200   2   = 19900 sets; we have to prove
that at least one set is empty.

                                        1
   For each problem, there are at most 80 students who did not solve it.
From these students at most 80  2   = 3160 pairs can be selected, so the
problem can belong to at most 3160 sets. The 6 problems together can
belong to at most 6 · 3160 = 18960 sets.
   Hence, at least 19900 − 18960 = 940 sets must be empty.
Problem 3. For each n ≥ 1 let
                                   ∞                        ∞
                                   X kn                     X             kn
                            an =              ,     bn =          (−1)k      .
                                         k!                               k!
                                   k=0                      k=0
    Show that an · bn is an integer.
Solution. We prove by induction on n that a n /e and bn e are integers, we
prove this for n = 0 as well. (For n = 0, the term 0 0 in the definition of the
sequences must be replaced by 1.)
    From the power series of ex , an = e1 = e and bn = e−1 = 1/e.
    Suppose that for some n ≥ 0, a0 , a1 , . . . , an and b0 , b1 , ...bn are all multi-
pliers of e and 1/e, respectively. Then, by the binomial theorem,

                    n                         ∞                         ∞ X
                                                                          n   m
                    X (k + 1)n+1              X (k + 1)n                X    n k
        an+1 =                           =                         =                       =
                          (k + 1)!                      k!                        m   k!
                    k=0                       k=0                       k=0 m=0

                                  ∞
                             n  X       n  
                            X   n   km   X   n
                          =            =        am
                            m=0
                                m   k!   m=0
                                             m
                                              k=0
    and similarly
                     n                                            ∞
                     X                (k + 1)n+1    X       (k + 1)n
            bn+1 =          (−1)k+1              =−   (−1)k          =
                                       (k + 1)!                k!
                      k=0                                         k=0

      ∞             n   m                            ∞
                                                  n  X                           n  
      X             X  n k                        X  n                      km    X   n
 =−         (−1)k                     =−                           (−1)k       =−        bm .
                    m=0
                            m   k!            m=0
                                                        m                   k!    m=0
                                                                                      m
      k=0                                                    k=0

   The numbers an+1 and bn+1 are expressed as linear combinations of the
previous elements with integer coefficients which finishes the proof.
Problem 4. In the tetrahedron OABC, let ∠BOC = α, ∠COA = β and
∠AOB = γ. Let σ be the angle between the faces OAB and OAC, and let
τ be the angle between the faces OBA and OBC. Prove that

                                   γ > β · cos σ + α · cos τ.

Solution. We can assume OA = OB = OC = 1. Intersect the unit sphere
with center O with the angle domains AOB, BOC and COA; the intersec-
tions are “slices” and their areas are 12 γ, 21 α and 12 β, respectively.

                                                    2
     Now project the slices AOC and COB to the plane OAB. Denote by
C 0 the projection of vertex C, and denote by A 0 and B 0 the reflections of
vertices A and B with center O, respectively. By the projection, OC 0 < 1.
     The projections of arcs AC and BC are segments of ellipses with long
axes AA0 and BB 0 , respectively. (The ellipses can be degenerate if σ or τ
is right angle.) The two ellipses intersect each other in 4 points; both half
ellipses connecting A and A0 intersect both half ellipses connecting B and
B 0 . There exist no more intersection, because two different conics cannot
have more than 4 common points.
     The signed areas of the projections of slices AOC and COB are 12 α·cos τ
and 21 β · cos σ, respectively. The statement says thet the sum of these signed
areas is less than the area of slice BOA.




    There are three significantly different cases with respect to the signs
of cos σ and cos τ (see Figure). If both signs are positive (case (a)), then
the projections of slices OAC and OBC are subsets of slice OBC without
common interior point, and they do not cover the whole slice OBC; this
implies the statement. In cases (b) and (c) where at least one of the signs
is negative, projections with positive sign are subsets of the slice OBC, so
the statement is obvious again.
Problem 5. Let A be an n × n matrix with complex entries and suppose
that n > 1. Prove that
                                                                 −1
           AA = In ⇐⇒ ∃ S ∈ GLn (C)         such that   A = SS        .

(If A = [aij ] then A = [aij ], where aij is the complex conjugate of aij ;
GLn (C) denotes the set of all n×n invertible matrices with complex entries,
and In is the identity matrix.)
                                                                          −1
Solution.    The direction ⇐ is trivial, since if A = SS , then
          −1
AA = SS · SS −1 = In .
    For the direction ⇒, we must prove that there exists an invertible matrix
S such that AS = S.
    Let w be an arbitrary complex number which is not 0. Choosing
S = wA + wIn , we have AS = A(wA + wIn ) = wIn + wA = S. If S is
singular, then w1 S = A − (w/w)In is singular as well, so w/w is an eigen-
value of A. Since A has finitely many eigenvalues and w/w can be any
complex number on the unit circle, there exist such w that S is invertible.

                                      3
Problem 6. Let f : Rn → R be a convex function whose gradient ∇f =

  ∂f            ∂f                            n
  ∂x1 , . . . , ∂xn exists at every point of R and satisfies the condition

        ∃L > 0    ∀x1 , x2 ∈ Rn      k∇f (x1 ) − ∇f (x2 )k ≤ Lkx1 − x2 k.
   Prove that

 ∀x1 , x2 ∈ Rn   k∇f (x1 ) − ∇f (x2 )k2 ≤ Lh∇f (x1 ) − ∇f (x2 ), x1 − x2 i. (1)

   In this formula ha, bi denotes the scalar product of the vectors a and b.
Solution. Let g(x) = f (x)−f (x1 )−h∇f (x1 ), x−x1 i. It is obvious that g has
the same properties. Moreover, g(x 1 ) = ∇g(x1 ) = 0 and, due to convexity,
g has 0 as the absolute minimum at x1 . Next we prove that
                                           1
                            g(x2 ) ≥         k∇g(x2 )k2 .                              (2)
                                          2L
Let y0 = x2 − L1 k∇g(x2 )k and y(t) = y0 + t(x2 − y0 ). Then
                                     Z 1
                 g(x2 ) = g(y0 ) +         h∇g(y(t)), x2 − y0 i dt =
                                      0
                                       Z 1
    = g(y0 ) + h∇g(x2 ), x2 − y0 i −           h∇g(x2 ) − ∇g(y(t)), x2 − y0 i dt ≥
                                           0
                               Z 1
              1
      ≥0+       k∇g(x2 )k2 −         k∇g(x2 ) − ∇g(y(t))k · kx2 − y0 k dt ≥
              L                 0
                                                   Z 1
               1
              ≥ k∇g(x2 )k2 − kx2 − y0 k                  Lkx2 − g(y)k dt =
               L                                    0
                                                   Z 1
              1                                                    1
          =     k∇g(x2 )k2 − Lkx2 − y0 k2                t dt =      k∇g(x2 )k2 .
              L                                     0             2L
   Substituting the definition of g into (2), we obtain
                                                          1
      f (x2 ) − f (x1 ) − h∇f (x1 ), x2 − x1 i ≥            k∇f (x2 ) − ∇f (x1 )k2 ,
                                                         2L

   k∇f (x2 ) − ∇f (x1 )k2 ≤ 2Lh∇f (x1 ), x1 − x2 i + 2L(f (x2 ) − f (x1 )).            (3)

Exchanging variables x1 and x2 , we have

   k∇f (x2 ) − ∇f (x1 )k2 ≤ 2Lh∇f (x2 ), x2 − x1 i + 2L(f (x1 ) − f (x2 )).            (4)

The statement (1) is the average of (3) and (4).




                                               4
