                                       IMC 2022
                            Second Day, August 4, 2022



Problem 5. We colour all the sides and diagonals of a regular polygon P with 43 vertices
either red or blue in such a way that every vertex is an endpoint of 20 red segments and 22
blue segments. A triangle formed by vertices of P is called monochromatic if all of its sides
have the same colour. Suppose that there are 2022 blue monochromatic triangles. How many
red monochromatic triangles are there?
                                                                                 (10 points)




Problem 6. Let p > 2 be a prime number. Prove that there is a permutation (x1 , x2 , ..., xp−1 )
of the numbers (1, 2, ..., p − 1) such that
                             x1 x2 + x2 x3 + ... + xp−2 xp−1 ≡ 2 (mod p).
                                                                                (10 points)




Problem 7. Let A1 , A2 , . . . , Ak be n × n idempotent complex matrices such that

                                 Ai Aj = −Aj Ai   for all i ̸= j.

Prove that at least one of the given matrices has rank ≤ nk .
(A matrix A is called idempotent if A2 = A.)
                                                                                     (10 points)



Problem 8. Let n, k ≥ 3 be integers, and let S be a circle. Let n blue points and k red
points be chosen uniformly and independently at random on the circle S. Denote by F the
intersection of the convex hull of the red points and the convex hull of the blue points. Let m
be the number of vertices of the convex polygon F (in particular, m = 0 when F is empty).
Find the expected value of m.
                                                                                     (10 points)
