                                          IMC 2025
                                First Day, July 30, 2025



Problem 1. Let P ∈ R[x] be a polynomial with real coefficients, and suppose deg(P ) ≥ 2. For
every x ∈ R, let ℓx ⊂ R2 denote the line tangent to the graph of P at the point (x, P (x)).
                                                        [
   (a) Suppose that the degree of P is odd. Show that      ℓx = R2 .
                                                             x∈R
   (b) Does there exist a polynomial of even degree for which the above equality still holds?

                                                                                        (10 points)


Problem
R1          2. Let f : R → R be a twice continuously differentiable function, and suppose that
 −1
    f (x) dx = 0 and f (1) = f (−1) = 1. Prove that
                                     Z 1
                                                   2
                                         (f ′′ (x)) dx ≥ 15,
                                         −1

and find all such functions for which equality holds.

                                                                                        (10 points)


Problem 3. Denote by S the set of all real symmetric 2025 × 2025 matrices of rank 1 whose
entries take values −1 or +1. Let A, B ∈ S be matrices chosen independently uniformly at
random. Find the probability that A and B commute, i.e. AB = BA.

                                                                                        (10 points)


Problem 4. Let a be an even positive integer. Find all real numbers x such that
                             j√              k
                                b + x · ba−1 = ba + ⌊x/a⌋
                              a a
                                                                                                 (1)

holds for every positive integer b.
(Here ⌊x⌋ denotes the largest integer that is no greater than x.)

                                                                                        (10 points)


Problem 5. For a positive integer n, let [n] = {1, 2, . . . , n}. Denote by Sn the set of all
bijections from [n] to [n], and let Tn be the set of all maps from [n] to [n]. Define the order
ord(τ ) of a map τ ∈ Tn as the number of distinct maps in the set {τ, τ ◦ τ, τ ◦ τ ◦ τ, . . .} where
◦ denotes composition. Finally, let
                         f (n) = max ord(τ ) and g(n) = max ord(τ ).
                                 τ ∈Sn                         τ ∈Tn

Prove that g(n) < f (n) + n0.501 for sufficiently large n.

                                                                                        (10 points)
