      12th International Mathematics Competition for University
                              Students
                             Blagoevgrad, July 22 - July 28, 2005
                                                    Second Day


Problem 1. Let f (x) = x2 + bx + c, where b and c are real numbers, and let

                                         M = {x ∈ R : |f (x)| < 1}.

Clearly the set M is either empty or consists of disjoint open intervals. Denote the sum of
their lengths by |M |. Prove that               √
                                        |M | ≤ 2 2.
                               2                     2
Solution. Write f (x) = x + 2b + d where d = c − b4 . The absolute minimum of f is d.
   If d ≥ 1 then f (x) ≥ 1 for all x, M = ∅ and |M | = 0.
   If −1 < d < 1 then f (x) > −1 for all x,
                                  2
                                                                  √
                         
                                b                           b
                   −1 < x +          + d < 1 ⇐⇒         x+     < 1−d
                                2                           2
so
                                              b √        b √
                                                                
                                 M=          − − 1 − d, − + 1 − d
                                              2          2
and                                                √         √
                                           |M | = 2 1 − d < 2 2.
     If d ≤ −1 then
                       2
                      b                                      p                       b  p
             −1 < x +      +d<1                     ⇐⇒           |d| − 1 < x +         < |d| + 1
                      2                                                              2
so                                p             p         p         p       
                     M= −             |d| + 1, − |d| − 1 ∪   |d| − 1, |d| + 1
and
            p               p             (|d| + 1) − (|d| − 1)         2        √
 |M | = 2        |d| + 1 −      |d| − 1 = 2 p           p         ≤ 2√      √   = 2 2.
                                              |d| + 1 + |d| − 1        1+1+ 1−0

Problem 2. Let f : R → R be a function such that (f (x))n is a polynomial for every
n = 2, 3, . . .. Does it follow that f is a polynomial?
Solution 1. Yes, it is even enough to assume that f 2 and f 3 are polynomials.
   Let p = f 2 and q = f 3 . Write these polynomials in the form of

                             p = a · pa11 · . . . · pakk ,   q = b · q1b1 · . . . · qlbl ,
where a, b ∈ R, a1 , . . . , ak , b1 , . . . bl are positive integers and p1 , . . . , pk , q1 , . . . , ql are irre-
ducible polynomials with leading coefficients 1. For p3 = q 2 and the factorisation of p3 = q 2
is unique we get that a3 = b2 , k = l and for some (i1 , . . . , ik ) permutation of (1, . . . , k) we
have p1 = qi1 , . . . , pk = qik and 3a1 = 2bi1 , . . . , 3ak = 2bik . Hence b1 , . . . , bl are divisible by
                   b /3         b /3
3 let r = b1/3 · q11 · . . . · ql l be a polynomial. Since r3 = q = f 3 we have f = r.
                                                                                   3
Solution 2. Let pq be the simplest form of the rational function ff 2 . Then the simplest form
                                               3 2
                 p2                     p2
of its square is q2 . On the other hand q2 = ff 2 = f 2 is a polynomial therefore q must
                                     3
be a constant and so f = ff 2 = pq is a polynomial.

Problem 3. In the linear space of all real n × n matrices, find the maximum possible
dimension of a linear subspace V such that

                                          ∀X, Y ∈ V      trace(XY ) = 0.

(The trace of a matrix is the sum of the diagonal entries.)
Solution. If A is a nonzero symmetric matrix, then trace(A2 ) = trace(At A) is the sum of
the squared entries of A which is positive. So V cannot contain any symmetric matrix but
0.
   Denote by S the linear space of all real n × n symmetric matrices; dim V = n(n+1)  2
                                                                                         .
                                                 2                   2   n(n+1)   n(n−1)
Since V ∩ S = {0}, we have dim V + dim S ≤ n and thus dim V ≤ n − 2 = 2 .
   The space of strictly upper triangular matrices has dimension n(n−1)
                                                                     2
                                                                         and satisfies the
condition of the problem.
   Therefore the maximum dimension of V is n(n−1)  2
                                                     .

Problem 4. Prove that if f : R → R is three times differentiable, then there exists a real
number ξ ∈ (−1, 1) such that

                                         f 000 (ξ)   f (1) − f (−1)
                                                   =                − f 0 (0).
                                             6              2

Solution 1. Let
                  f (−1) 2                          f (1) 2
     g(x) = −           x (x − 1) − f (0)(x2 − 1) +      x (x + 1) − f 0 (0)x(x − 1)(x + 1).
                     2                                2
It is easy to check that g(±1) = f (±1), g(0) = f (0) and g 0 (0) = f 0 (0).
     Apply Rolle’s theorem for the function h(x) = f (x) − g(x) and its derivatives. Since
h(−1) = h(0) = h(1) = 0, there exist η ∈ (−1, 0) and ϑ ∈ (0, 1) such that h0 (η) =
h0 (ϑ) = 0. We also have h0 (0) = 0, so there exist % ∈ (η, 0) and σ ∈ (0, ϑ) such that
h00 (%) = h00 (σ) = 0. Finally, there exists a ξ ∈ (%, σ) ⊂ (−1, 1) where h000 (ξ) = 0. Then

                               f (−1)                   f (1)                     f (1) − f (−1)
   f 000 (ξ) = g 000 (ξ) = −          · 6 − f (0) · 0 +       · 6 − f 0 (0) · 6 =                − f 0 (0).
                                  2                       2                              2
                           f (1) − f (−1)
Solution 2. The expression                − f 0 (0) is the divided difference f [−1, 0, 0, 1] and
                                  2
                                                                f 000 (ξ)
there exists a number ξ ∈ (−1, 1) such that f [−1, 0, 0, 1] =             .
                                                                   3!
Problem 5. Find all r > 0 such that whenever f : R2 → R is a differentiable function
such that |grad f (0, 0)| = 1 and |grad f (u) − grad f (v)| ≤ |u − v| for all u, v ∈ R2 , then
the maximum of f on the disk {u ∈ R2 : |u| ≤ r} is attained at exactly one point.
(grad f (u) = (∂1√f (u), ∂2 f (u)) is the gradient vector of f at the point u. For a vector
u = (a, b), |u| = a2 + b2 .)
                                                                         x2 y 2
Solution. To get an upper bound for r, set f (x, y) = x −                   + . This function satisfies
                                                                         2    2
the conditions, since grad f (x, y) = (1 − x, y), grad f (0, 0) = (1, 0) and |grad f (x1 , y1 ) −
grad f (x2 , y2 )| = |(x2 − x1 , y1 − y2 )| = |(x1 , y1 ) − (x2 , y2 )|.
    In the disk Dr = {(x, y) : x2 + y 2 ≤ r2 }
                                                             2
                                      x2 + y 2                             r2 1
                                                   
                                                           1         1
                          f (x, y) =            − x−              + ≤        + .
                                           2               2         4     2    4
                                                                                          q         
                                                    2
If r > 21 then the absolute maximum is r2 + 14 , attained at the points 12 , ± r2 − 14 .
Therefore, it is necessary that r ≤ 12 because if r > 21 then the maximum is attained twice.
    Suppose now that r ≤ 1/2 and that f attains its maximum on Dr at u, v, u 6= v. Since
|grad f (z) − grad f (0)| ≤ r, |grad f (z)| ≥ 1 − r > 0 for all z ∈ Dr . Hence f may attain its
maximum only at the boundary of Dr , so we must have |u| = |v| = r and grad f (u) = au
and grad f (v) = bv, where a, b ≥ 0. Since au = grad f (u) and bv = grad f (v) belong
to the disk D with centre grad f (0) and radius r, they do not belong to the interior of
Dr . Hence |grad f (u) − grad f (v)| = |au − bv| ≥ |u − v| and this inequality is strict
since D ∩ Dr contains no more than one point. But this contradicts the assumption that
|grad f (u) − grad f (v)| ≤ |u − v|. So all r ≤ 12 satisfies the condition.
                                                                           √
Problem6. Prove  that         and q are rational numbers and r = p + q 7, then there exists
                          if p 
            a b            1 0
a matrix           6= ±           with integer entries and with ad − bc = 1 such that
            c d            0 1
                                             ar + b
                                                    = r.
                                             cr + d
Solution. First consider the case when q = 0 and r is rational. Choose a positive integer t
such that r2 t is an integer and set
                                                      
                                  a b      1 + rt −r2 t
                                       =                   .
                                  c d        t    1 − rt
Then
                                                         (1 + rt)r − r2 t
                            
                         a b                    ar + b
                     det       = 1 and                 =                  = r.
                         c d                    cr + d    tr + (1 − rt)
    Now assume q 6= 0. Let the minimal
                                    √     polynomial of r in Z[x] be ux2 + vx + w. The other
root of this polynomial is r = p−q 7, so v = −u(r+r) = −2up and w = urr = u(p2 −7q 2 ).
The discriminant is v 2 − 4uw = 7 · (2uq)2 . The left-hand side is an integer, implying that
also ∆ = 2uq is an integer.
    The equation ar+b
                   cr+d
                        = r is equivalent to cr2 + (d − a)r − b = 0. This must be a multiple
of the minimal polynomial, so we need

                                 c = ut,    d − a = vt,    −b = wt

for some integer t 6= 0. Putting together these equalities with ad − bc = 1 we obtain that

                 (a + d)2 = (a − d)2 + 4ad = 4 + (v 2 − 4uw)t2 = 4 + 7∆2 t2 .

Therefore 4 + 7∆2 t2 must be a perfect square. Introducing s = a + d, we need an integer
solution (s, t) for the Diophantine equation

                                           s2 − 7∆2 t2 = 4                                        (1)

such that t 6= 0.
    The numbers s and t will be even. Then a + d = s and d − a = vt will be even as well
and a and d √   will be really integers.
                                 √                                                 √            √
        (8±3 7)n =
    Let √               √k n ±ln   7 for each integer n. Then k 2
                                                                n −7ln
                                                                      2
                                                                        = (k n +ln   7)(k n −ln   7) =
((8 + 3 7)n (8 − 3 7))n = 1 and the sequence (ln ) also satisfies the linear recurrence
ln+1 = 16ln − ln−1 . Consider the residue of ln modulo ∆. There are ∆2 possible residue
pairs for (ln , ln+1 ) so some are the same. Starting from such two positions, the recurrence
shows that the sequence of residues is periodic in both directions. Then there are infinitely
many indices such that ln ≡ l0 = 0 (mod ∆).
    Taking such an index n, we can set s = 2kn and t = 2ln /∆.
Remarks. 1. It is well-known that if D > 0 is not a perfect square then the Pell-like
Diophantine equation
                                             x2 − Dy 2 = 1
has infinitely many solutions. Using this fact the solution can be generalized to all quadratic
algebraic numbers.
2. It is also known that the continued fraction of a real number r is periodic from a certain
point if and only if r is a root of a quadratic equation. This fact can lead to another
solution.
