                                    IMC 2021 Online
                              Second Day, August 4, 2021
                                            Solutions


Problem 5. Let A be a real n × n matrix and suppose that for every positive integer m there exists
a real symmetric matrix B such that

                                         2021B = Am + B 2 .

Prove that ∣ det A∣ ≤ 1.
            (proposed by Rafael Filipe dos Santos, Instituto Militar de Engenharia, Rio de Janeiro)

Hint: The determinant is the product of the eigenvalues.

Solution. Let Bm be the corresponding matrix B depending on m:

                                        2021Bm = Am + Bm
                                                       2
                                                         .

    For m = 1, we obtain A = 2021B1 − B12 . Since B1 is real and symmetric, so is A. Thus A is
diagonalizable and all eigenvalues of A are real.
    Now fix a positive integer m and let λ be any real eigenvalue of A. Considering the diagonal form
of both A and Bm , we know that there exists a real eigenvalue µ of Bm such that

                              2021µ = λm + µ2 ⇒ µ2 − 2021µ + λm = 0.

    The last equation is a second degree equation with a real root. Therefore, the discriminant is
non-negative:
                                                               20212
                                       20212 − 4λm ≥ 0 ⇒ λm ≤        .
                                                                 4
    If ∣λ∣ > 1, letting m even sufficiently large we reach a contradiction. Thus ∣λ∣ ≤ 1.
    Finally, since det A is the product of the eigenvalues of A and each of them has absolute value
less then or equal to 1, we get ∣ det A∣ ≤ 1 as desired.

Solution. Different solution can be found in paper s2002

Problem 6. For a prime number p, let GL2 (Z/pZ) be the group of invertible 2 × 2 matrices of
residues modulo p, and let Sp be the symmetric group (the group of all permutations) on p elements.
Show that there is no injective group homomorphism ϕ ∶ GL2 (Z/pZ) → Sp .
                                          (proposed by Thiago Landim, Sorbonne University, Paris)

Hint: First find what the monomorphism must do with elements of order p.

Solution. For p = 2, just note that GL2 (Z/2Z) has more than 2 = ∣S2 ∣ elements.
   From now on, let p be an odd prime and suppose that there exists such a homomorphism.
   The matrix
                                                 1 1
                                            A=(      )
                                                 0 1



                                                 1
has order p and commutes with the matrix

                                                         −1 0
                                               B=(            )
                                                         0 −1

of order 2, hence AB has order 2p. But there is no permutation in Sp of order 2p since only p-cycles
have order divisible by p, and their order is exactly p.

Problem 7. Let D ⊆ C be an open set containing the closed unit disk {z ∶ ∣z∣ ≤ 1}. Let f ∶ D → C
be a holomorphic function, and let p(z) be a monic polynomial. Prove that

                                          ∣f (0)∣ ≤ max ∣f (z)p(z)∣.
                                                      ∣z∣=1


                                                                                (proposed by Lars Hörmander)

Hint: Apply the maximum principle or the Cauchy formula to a suitable function f (z)q(z).

Solution.
   Let q(z) = z n ⋅ p (1/z), or more explicitly, if

                                      p(z) = z n + an−1 z n−1 + ⋅ ⋅ ⋅ + a0 ,

let
                                       q(z) = 1 + an−1 z + ⋅ ⋅ ⋅ + a0 z n .
Note that for ∣z∣ = 1 we have 1/z = z and hence ∣q(z)∣ = ∣p(z)∣. Hence by the maximum principle or
the Cauchy formula for the product of f and q, it follows that

                          ∣f (0)∣ = ∣f (0)q(0)∣ ≤ max ∣f (z)q(z)∣ = max ∣f (z)p(z)∣.
                                                 ∣z∣=1                  ∣z∣=1




Problem 8. Let n be a positive integer. At most how many distinct unit vectors can be selected in
Rn such that from any three of them, at least two are orthogonal?
                 (proposed by Alexander Polyanskii, Moscow Institute of Physics and Technology;
                               based on results of Paul Erdős and Moshe Rosenfeld)
Hint: Play with the Gram matrix of these vectors.

Solution 1. 2n is the maximal number.
    An example of 2n vectors in the set is given by a basis and its opposite vectors. In the rest of
the text we prove that it is impossible to have 2n + 1 vectors in the set.
    Consider the Gram matrix A with entries aij = ei ⋅ ej . Its rank is at most n, its eigenvalues are
real and non-negative. Put B = A − I2n+1 , this is the same matrix, but with zeros on the diagonal.
The eigenvalues of B are real, greater or equal to −1, and the multiplicity of −1 is at least n + 1.
    The matrix C = B 3 has the following diagonal entries

                                            cii = ∑ aij ajk aki .
                                                  i≠j≠k≠i

The problem statement implies that in every summand of this expression at least one factor is zero.
Hence tr C = 0. Let x1 , . . . , xm be the positive eigenvalues of B, their number is m ≤ n as noted
above. From tr B = tr C we deduce (taking into account that the eigenvalues between −1 and 0
satisfy λ3 ≥ λ):
                                        x1 + ⋅ ⋅ ⋅ + xm ≥ x31 + ⋅ ⋅ ⋅ + x3m

                                                         2
Applying tr C = 0 once again and noting that C has eigenvalue −1 of multiplicity at least n + 1, we
obtain
                                      x31 + ⋅ ⋅ ⋅ + x3m ≥ n + 1.
It also follows that
                                   (x1 + ⋅ ⋅ ⋅ + xm ) ≥ (x31 + ⋅ ⋅ ⋅ + x3m ) (n + 1)2 .
                                                     3


By Hölder’s inequality, we obtain

                                      (x31 + ⋅ ⋅ ⋅ + x3m ) m2 ≥ (x1 + ⋅ ⋅ ⋅ + xm )3 ,

which is a contradiction with m ≤ n.

Solution 2. Let Pi denote the projection onto i-th vector, i = 1, . . . , N . Then our relation reads as
tr (Pi Pj Pk ) = 0 for distinct i, j, k. Consider the operator Q = ∑N
                                                                    i=1 Pi , it is non-negative definite, let
t1 , . . . , tn be its eigenvalues, ∑ ti = tr Q = N . We get

                      ∑ t3i = tr Q3 = N + 6 ∑ tr Pi Pj = N + 3(tr Q2 − N ) = 3 ∑ t2i − 2N
                                              i<j


(we used the obvious identities like trPi Pj Pi = tr Pi2 Pj = tr Pi Pj ). But (ti − 2)2 (ti + 1) = t3i − 3t2i + 4 ⩾ 0,
thus −2N = ∑ t3i − 3t2i ⩾ −4n and N ⩽ 2n.




                                                            3
