                           IMC 2012, Blagoevgrad, Bulgaria
                                          Day 1, July 28, 2012



Problem 1. Let A and B be real symmetric matrices with all eigenvalues strictly greater than 1. Let
λ be a real eigenvalue of matrix AB. Prove that |λ| > 1.
                                                  (Proposed by Pavel Kozhevnikov, MIPT, Moscow)

Solution. The transforms given by A and B strictly increase the length of every nonzero vector, this
can be seen easily in a basis where the matrix is diagonal with entries greater than 1 in the diagonal.
Hence their product AB also strictly increases the length of any nonzero vector, and therefore its real
eigenvalues are all greater than 1 or less than −1.


Problem 2. Let f : R → R be a twice differentiable function. Suppose f (0) = 0. Prove that there
exists ξ ∈ (−π/2, π/2) such that
                                 f 00 (ξ) = f (ξ)(1 + 2 tan2 ξ).
                             (Proposed by Karen Keryan, Yerevan State University, Yerevan, Armenia )

Solution. Let g(x) = f (x) cos x. Since g(−π/2) = g(0) = g(π/2) = 0, by Rolle’s theorem there exist
some ξ1 ∈ (−π/2, 0) and ξ2 ∈ (0, π/2) such that

                                               g 0 (ξ1 ) = g 0 (ξ2 ) = 0.

Now consider the function
                                        g 0 (x)   f 0 (x) cos x − f (x) sin x
                                  h(x) =        =                             .
                                        cos2 x               cos2 x
We have h(ξ1 ) = h(ξ2 ) = 0, so by Rolle’s theorem there exist ξ ∈ (ξ1 , ξ2 ) for which

                         g 00 (ξ) cos2 ξ + 2 cos ξ sin ξg 0 (ξ)
          0 = h0 (ξ) =                                          =
                                         cos4 ξ
              (f 00 (ξ) cos ξ − 2f 0 (ξ) sin ξ − f (ξ) cos ξ) cos ξ + 2 sin ξ(f 0 (ξ) cos ξ − f (ξ) sin ξ)
            =                                                                                              =
                                                            cos3 ξ
              f 00 (ξ) cos2 ξ − f (ξ)(cos2 ξ + 2 sin2 ξ)           1
            =                          3
                                                              =       (f 00 (ξ) − f (ξ)(1 + 2 tan2 ξ)).
                                   cos ξ                        cos ξ
The last yields the desired equality.


Problem 3. There are 2n students in a school (n ∈ N, n ≥ 2). Each week n students go on a trip.
After several trips the following condition was fulfilled: every two students were together on at least
one trip. What is the minimum number of trips needed for this to happen?
                                                        (Proposed by Oleksandr Rybak, Kiev, Ukraine)

Solution. We prove that for any n ≥ 2 the answer is 6.
    First we show that less than 6 trips is not sufficient. In that case the total quantity of students in
all trips would not exceed 5n. A student meets n − 1 other students in each trip, so he or she takes
part on at least 3 excursions to meet all of his or her 2n − 1 schoolmates. Hence the total quantity of
students during the trips is not less then 6n which is impossible.
    Now let’s build an example for 6 trips.


                                                           1
    If n is even, we may divide 2n students into equal groups A, B, C, D. Then we may organize the
trips with groups (A, B), (C, D), (A, C), (B, D), (A, D) and (B, C), respectively.
    If n is odd and divisible by 3, we may divide all students into equal groups E, F , G, H, I, J.
Then the members of trips may be the following: (E, F, G), (E, F, H), (G, H, I), (G, H, J), (E, I, J),
(F, I, J).
    In the remaining cases let n = 2x+3y be, where x and y are natural numbers. Let’s form the groups
A, B, C, D of x students each, and E, F , G, H, I, J of y students each. Then we apply the previous
cases and organize the following trips: (A, B, E, F, G), (C, D, E, F, H), (A, C, G, H, I), (B, D, G, H, J),
(A, D, E, I, J), (B, C, F, I, J).


                                                                                            n                n
                                                                                                                   x2i
                                                                                            P                P
Problem 4. Let n ≥ 3 and let x1 , . . . , xn be nonnegative real numbers. Define A =              xi , B =
                                                                                            i=1              i=1
          n
                x3i . Prove that
          P
and C =
          i=1


                                   (n + 1)A2 B + (n − 2)B 2 ≥ A4 + (2n − 2)AC.

                                                      (Proposed by Géza Kós, Eötvös University, Budapest)

Solution. Let
                  n
                 Y                              A2 − B n−2 A3 − 3AB + 2C n−3
          p(X) =     (X − xi ) = X n − AX n−1 +       X   −             X    + ....
                 i=1
                                                   2             6

The (n − 3)th derivative of p has three nonnegative real roots 0 ≤ u ≤ v ≤ w. Hence,

      6 (n−3)             3A 2 3(A2 − B)     A3 − 3AB + 2C
         p    (X) = X 3 −    X +          X−                 = (X − u)(X − v)(X − w),
      n!                   n     n(n − 1)    n(n − 1)(n − 2)
so
                            3A                        3(A2 − B)                  A3 − 3AB + 2C
           u+v+w =             ,     uv + vw + wu =               and uvw =                      .
                             n                         n(n − 1)                  n(n − 1)(n − 2)
From these we can see that
                  n2 (n − 1)2 (n − 2)
                                      (n + 1)A2 B + (n − 2)B 2 − A4 − (2n − 2)AC = . . . =
                                                                                
                           9
= u2 v 2 + v 2 w2 + w2 u2 − uvw(u + v + w) = uv(u − w)(v − w) + vw(v − u)(w − u) + wu(w − v)(u − v) ≥
                               ≥ 0 + uw(v − u)(w − v) + wu(w − v)(u − v) = 0.



Problem 5. Does
              P∞ there exist a sequence (an ) of complex numbers such that for every positive integer
                    p
p we have that n=1 an converges if and only if p is not a prime?
                                             (Proposed by Tomáš Bárta, Charles University, Prague)

Solution. The answer is YES. We prove a more general statement; suppose that N = C ∪ D is an
                                                                                  ∞
P∞ p decomposition of N into two disjoint sets. Then there exists a sequence (an )n=1 such that
arbitrary
  n=1 an is convergent for p ∈ C and divergent for p ∈ D.
   Define Ck = C ∩ [1, k] and Dk ∩ [1, k].




                                                        2
Lemma. For every positive integer k there exists a positive integer Nk and a sequence Xk = (xk,1 , . . . , xk,Nk )
of complex numbers with the following properties:
                                   Nk
                                   X
 (a) For p ∈ Dk , we have                 xpk,j ≥ 1.
                                   j=1

                                 Nk                                      m
                                 X                                       X                1
 (b) For p ∈ Ck , we have                xpk,j = 0; moreover,                  xpk,j ≤      holds for 1 ≤ m ≤ Nk .
                                 j=1                                     j=1
                                                                                          k

Proof. First we find some complex numbers z1 . . . , zk with
                                       k
                                                  (
                                      X              0 p ∈ Ck
                                          zjp =                                                                                 (1)
                                      j=1
                                                     1 p ∈ Dk

As is well-known, this system of equations is equivalent to another system σν (z1 , . . . , zk ) = wν (ν =
1, 2, . . . , k) where σν is the νth elementary symmetric polynomial, and the constants wν are uniquely
determined by the Newton-Waring-Girard formulas. Then the numbers z1 , . . . , zk are the roots of the
polynomial z k − w1 z k−1 + − . . . + (−1)k wk in some order.
    Now let                                    &                   '
                                                            Xm
                                          M=        max        zjp
                                                             1≤m≤k, p∈Ck
                                                                                  j=1
                       k                                                                     z1    z2            zk
and let Nk = k·(kM ) . We define the numbers xk,1 . . . , xk,Nk by repeating the sequence ( kM  , kM  , . . . , kM  )
                           z
(kM )k times, i.e. xk,` = kMj if ` ≡ j (mod k). Then we have
                                   Nk                          k                                 k
                                                                         zj p
                                   X                           X                                 X
                                          xpk,j = (kM )k                kM
                                                                               = (kM )k−p          zjp ;
                                    j=1                          j=1                              j=1

then from (1) the properties (a) and the first part of (b) follows immediately. For the second part of
(b), suppose that p ∈ Ck and 1 ≤ m ≤ Nk ; then m = kr + s with some integers r and 1 ≤ s ≤ k and
hence
                        m           kr    kr+s       s 
                       X    p
                                   X      X         X      zj p       M      1
                           xk,j =      +         =                ≤       p
                                                                            ≤ .
                       j=1         j=1   j=kr+1     j=1
                                                          kM         (kM )    k
The lemma is proved.
   Now let Sk = N1 . . . , Nk (we also define S0 = 0). Define the sequence (a) by simply concatenating
the sequences X1 , X2 , . . . :
                   (a1 , a2 , . . . ) = (x1,1 , . . . , x1,N1 , x2,1 , . . . , x2,N2 , . . . , xk,1 , . . . , xk,Nk , . . .);   (1)
                                            aSk +j = xk+1,j (1 ≤ j ≤ Nk+1 ).                                                    (2)
   If p ∈ D and k ≥ p then
                                                   Sk+1                 Nk+1
                                                   X                    X
                                                           apj    =            xpk+1,j ≥ 1;
                                                 j=Sk +1                 j=1
                                                  P p
By Cauchy’s convergence criterion it follows that  an is divergent.
   If p ∈ C and Su < n ≤ Su+1 with some u ≥ p then
                           n                       Nk
                                               u−1 X                    n−Su−1                n−Su−1
                           X                   X                            X                   X                   1
                                   apn    =                 xpk,j +               xpu,j   =             xpu,j ≤       .
                         j=Sp +1              k=p+1 j=1                     j=1                 j=1
                                                                                                                    u
                           ∞
                           X                                  ∞
                                                              X
Then it follows that               apn = 0, and thus                   apn = 0 is convergent.
                        n=Sp +1                               n=1



                                                                        3
