         13th International Mathematics Competition for University Students
                              Odessa, July 20-26, 2006
                                    Second Day

Problem 1. Let V be a convex polygon with n vertices.
    (a) Prove that if n is divisible by 3 then V can be triangulated (i.e. dissected into non-overlapping
triangles whose vertices are vertices of V ) so that each vertex of V is the vertex of an odd number
of triangles.
    (b) Prove that if n is not divisible by 3 then V can be triangulated so that there are exactly two
vertices that are the vertices of an even number of the triangles.
(20 points)
Solution. Apply induction on n. For the initial cases n = 3, 4, 5, chose the triangulations shown in
the Figure to prove the statement.
                             odd                 even                    odd                   odd



                                                                                odd                         odd




                   odd                     odd    odd                    even           even         even

    Now assume that the statement is true for some n = k and consider the case n = k + 3. Denote
the vertices of V by P1 , . . . , Pk+3 . Apply the induction hypothesis on the polygon P1 P2 . . . Pk ; in this
triangulation each of vertices P1 , . . . , Pk belong to an odd number of triangles, except two vertices
if n is not divisible by 3. Now add triangles P1 Pk Pk+2 , Pk Pk+1 Pk+2 and P1 Pk+2 Pk+3 . This way we
introduce two new triangles at vertices P1 and Pk so parity is preserved. The vertices Pk+1 , Pk+2 and
Pk+3 share an odd number of triangles. Therefore, the number of vertices shared by even number of
triangles remains the same as in polygon P1 P2 . . . Pk .
                                                        Pk       Pk −1
                                                                                Pk −2
                                     Pk +1



                                   Pk +2



                                     Pk +3
                                                                                P3
                                                    P1           P2


                                                                                                                  
Problem 2. Find all functions f : R −→ R such that for any real numbers a < b, the image f [a, b]
is a closed interval of length b − a.
(20 points)




                                                             1
Solution. The functions f (x) = x + c and f (x) = −x + c with some constant c obviously satisfy
the condition of the problem. We will prove now that these are the only functions with the desired
property.
    Let f be such a function. Then f clearly satisfies |f (x) − f (y)| ≤ |x − y| for all x, y; therefore, f
is continuous. Given x, y with x < y, let a, b ∈ [x, y] be such that f (a) is the maximum and f (b) is
the minimum of f on [x, y]. Then f ([x, y]) = [f (b), f (a)]; hence
                                     y − x = f (a) − f (b) ≤ |a − b| ≤ y − x
This implies {a, b} = {x, y}, and therefore f is a monotone function. Suppose f is increasing. Then
f (x) − f (y) = x − y implies f (x) − x = f (y) − y, which says that f (x) = x + c for some constant c.
Similarly, the case of a decreasing function f leads to f (x) = −x + c for some constant c.
Problem 3. Compare tan(sin x) and sin(tan x) for all x ∈ (0, π2 ).
(20 points)
Solution. Let f (x) = tan(sin x) − sin(tan x). Then
                                  cos x       cos(tan x)   cos3 x − cos(tan x) · cos2 (sin x)
                   f 0 (x) =                −            =
                               cos2 (sin x)     cos2 x            cos2 x · cos2 (tan x)
Let 0 < x < arctan π2 . It follows from the concavity of cosine on (0, π2 ) that
                                                                                              
       p3                 2
                                      1                                       tan x + 2 sin x
          cos(tan x) · cos (sin x) < [cos(tan x) + 2 cos(sin x)] ≤ cos                            < cos x ,
                                      3                                              3
                                    tan x+2 sin x 0                         q
                                                         1
                                                      = 3 cos2 x + 2 cos x ≥ 3 cos12 x · cos x · cos x = 1. This
                                                            1            
the last inequality follows from          3
proves that cos3 x−cos(tan x)·cos2 (sin x) > 0, so f 0 (x) > 0, so f increases on the interval [0, arctan π2 ].
To end the proof it is enough to notice that (recall that 4 + π 2 < 16)
                              h              π i              π/2           π
                          tan sin arctan              = tan p           > tan = 1 .
                                              2                     2
                                                              1 + π /4         4
This implies that if x ∈ [arctan π2 , π2 ] then tan(sin x) > 1 and therefore f (x) > 0.
Problem 4. Let v0 be the zero vector in Rn and let v1 , v2 , . . . , vn+1 ∈ Rn be such that the Euclidean
norm |vi − vj | is rational for every 0 ≤ i, j ≤ n + 1. Prove that v1 , . . . , vn+1 are linearly dependent
over the rationals.
(20 points)
Solution. By passing to a subspace we can assume that v1 , . . . , vn are linearly independent over the
reals. Then there exist λ1 , . . . , λn ∈ R satisfying
                                                             n
                                                             X
                                                  vn+1 =           λj v j
                                                             j=1

We shall prove that λj is rational for all j. From
                                      −2 hvi , vj i = |vi − vj |2 − |vi |2 − |vj |2
we get that hvi , vj i is rational for all i, j. Define A to be the rational n × n-matrix Aij = hvi , vj i,
w ∈ Qn to be the vector wi = hvi , vn+1 i, and λ ∈ Rn to be the vector (λi )i . Then,
                                                             n
                                                             X
                                            hvi , vn+1 i =         λj hvi , vj i
                                                             j=1

gives Aλ = w. Since v1 , . . . , vn are linearly independent, A is invertible. The entries of A−1 are
rationals, therefore λ = A−1 w ∈ Qn , and we are done.

                                                             2
Problem 5. Prove that there exists an infinite number of relatively prime pairs (m, n) of positive
integers such that the equation
                                        (x + m)3 = nx
has three distinct integer roots.
(20 points)
Solution. Substituting y = x + m, we can replace the equation by

                                              y 3 − ny + mn = 0.

Let two roots be u and v; the third one must be w = −(u + v) since the sum is 0. The roots must
also satisfy
                uv + uw + vw = −(u2 + uv + v 2 ) = −n,    i.e. u2 + uv + v 2 = n
and
                                       uvw = −uv(u + v) = mn.
So we need some integer pairs (u, v) such that uv(u + v) is divisible by u2 + uv + v 2 . Look for such
pairs in the form u = kp, v = kq. Then

                                   u2 + uv + v 2 = k 2 (p2 + pq + q 2 ),

and
                                           uv(u + v) = k 3 pq(p + q).
                                                                                       uv(u + v)
Chosing p, q such that they are coprime then setting k = p2 + pq + q 2 we have                      =
                                                                                      u2 + uv + v 2
p2 + pq + q 2 .
   Substituting back to the original quantites, we obtain the family of cases

                               n = (p2 + pq + q 2 )3 ,       m = p2 q + pq 2 ,

and the three roots are
                              x1 = p 3 ,       x2 = q 3 ,    x3 = −(p + q)3 .

Problem 6. Let Ai , Bi , Si (i = 1, 2, 3) be invertible real 2 × 2 matrices such that
   (1) not all Ai have a common real eigenvector;
   (2) Ai = Si−1 Bi Si for all i 
                                 = 1, 2,3;
                                   1 0
   (3) A1 A2 A3 = B1 B2 B3 =             .
                                   0 1
Prove that there is an invertible real 2 × 2 matrix S such that Ai = S −1 Bi S for all i = 1, 2, 3.
(20 points)
Solution. We note that the problem is trivial if Aj = λI for some j, so suppose this is not the case.
Consider then first the situation
                                 where some Aj , say A3 , has two distinct real eigenvalues.  We may
assume that A3 = B3 = λ µ by conjugating both sides. Let A2 = ( ac db ) and B2 = ac0 db0 . Then
                                                                                          0 0




                               a + d = Tr A2 = Tr B2 = a0 + d0
                aλ + dµ = Tr(A2 A3 ) = Tr A−1
                                           1  = Tr B1−1 = Tr(B2 B3 ) = a0 λ + d0 µ.


Hence a = a0 and d = d0 and so also bc = b0 c0 . Now we cannot have c = 0 or b = 0, for then (1, 0)> or
(0, 1)> would be a common eigenvector of all Aj . The matrix S = ( c c ) conjugates A2 = S −1 B2 S,
                                                                       0


and as S commutes with A3 = B3 , it follows that Aj = S −1 Bj S for all j.

                                                       3
    If the distinct eigenvalues of A3 = B3 are not real, we know from above that Aj = S −1 Bj S for
some S ∈ GL2 C unless all Aj have a common eigenvector over C. Even if they do, say Aj v = λj v,
by taking the conjugate square    root it follows that Aj ’s can be simultaneously diagonalized. If
A2 = ( a d ) and B2 = ac0 db0 , it follows as above that a = a0 , d = d0 and so b0 c0 = 0. Now B2
                             0 0


and B3 (and hence B1 too) have a common eigenvector over C so they too can be simultaneously
diagonalized. And so SAj = Bj S for some S ∈ GL2 C in either case. Let S0 = Re S and S1 = Im S.
By separating the real and imaginary components,     we are done if either S0 or S1 is invertible. If not,
                                                x 0
                                    −1
S0 may be conjugated to some T S0 T = y 0 , with (x, y) 6= (0, 0)> , and it follows that all Aj
                                                                >

have a common eigenvector T (0, 1)> , a contradiction.
    We are left with the case when no Aj has distinct eigenvalues; then these eigenvalues by necessity
are real. By conjugation and division by scalars we may assume that A3 = ( 1 1b ) and b 6= 0. By further
conjugation by upper-triangular matrices (which preserves the shape of A3 up to the value of b) we can  
                          0 u        2      2                                     −1 −1       −(b+v)/u 1
also assume that A2 = ( 1 v ). Here v = Tr A2 = 4 det A2 = −4u. Now A1 = A3 A2 =                 1/u       ,
and hence (b + v)2 /u2 = Tr2 A1 = 4 det A1 = −4/u. Comparing these two it follows that b = −2v.
What we have done is simultaneously reduced all Aj to matrices whose all entries depend on u and
v (= − det A2 and Tr A2 , respectively) only, but these themselves are invariant under similarity. So
Bj ’s can be simultaneously reduced to the very same matrices.




                                                     4
