           Solutions for problems in the
   9th International Mathematics Competition
             for University Students
                   Warsaw, July 19 - July 25, 2002
                                   First Day

Problem 1. A standard parabola is the graph of a quadratic polynomial
y = x2 + ax + b with leading coefficient 1. Three standard parabolas with
vertices V1 , V2 , V3 intersect pairwise at points A1 , A2 , A3 . Let A 7→ s (A) be
the reflection of the plane with respect to the x axis.
   Prove that standard parabolas with vertices s (A1 ), s (A2 ), s (A3 ) intersect
pairwise at the points s (V1 ), s (V2 ), s (V3 ).
Solution. First we show that the standard parabola with vertex V contains
point A if and only if the standard parabola with vertex s(A) contains point
s(V ).
     Let A = (a, b) and V = (v, w). The equation of the standard parabola
with vertex V = (v, w) is y = (x − v)2 + w, so it contains point A if and
only if b = (a − v)2 + w. Similarly, the equation of the parabola with vertex
s(A) = (a, −b) is y = (x − a)2 − b; it contains point s(V ) = (v, −w) if and
only if −w = (v − a)2 − b. The two conditions are equivalent.
     Now assume that the standard parabolas with vertices V1 and V2 , V1 and
V3 , V2 and V3 intersect each other at points A3 , A2 , A1 , respectively. Then, by
the statement above, the standard parabolas with vertices s(A1 ) and s(A2 ),
s(A1 ) and s(A3 ), s(A2 ) and s(A3 ) intersect each other at points V3 , V2 , V1 ,
respectively, because they contain these points.
Problem 2. Does there exist a continuously differentiable function f : R → R
such that for every x ∈ R we have f (x) > 0 and f 0 (x) = f (f (x))?
Solution. Assume that there exists such a function. Since f 0 (x) = f (f (x)) > 0,
the function is strictly monotone increasing.
    By the monotonity, f (x) > 0 implies f (f (x)) > f (0) for all x. Thus, f (0)
is a lower bound for f 0 (x), and for all x < 0 we have f (x) < f (0) + x · f (0) =
(1 + x)f (0). Hence, if x ≤ −1 then f (x) ≤ 0, contradicting the property
f (x) > 0.
    So such function does not exist.



                                        1
Problem 3. Let n be a positive integer and let
                           1
                   ak =    n ,
                                    bk = 2k−n ,        f or    k = 1, 2, . . . , n.
                           k

   Show that
                        a1 − b 1 a2 − b 2       an − b n
                                +         +···+          = 0.                                          (1)
                           1        2              n

Solution. Since k nk = n n−1
                            
                         k−1
                               for all k ≥ 1, (1) is equivalent to
               "                                          #
          2n       1             1                  1            21 22      2n
                   n−1 +        n−1 + · · · +
                                                 n−1
                                                        =         +   +···+ .                         (2)
          n         0            1                 n−1
                                                                 1   2      n

   We prove (2) by induction. For n = 1, both sides are equal to 2.
   Assume that (2) holds for some n. Let
                         "                             #
                      2n     1       1             1
                xn =        n−1 + n−1 + · · · + n−1
                                                     ;
                      n      0       1            n−1

then
                           n                              n−1
                                                                                      !        !
              2n+1 X 1      2n                            X         1         1
       xn+1 =        n
                        =                         1+               n +
                                                                            n
                                                                                         +1       =
              n+1    k
                           n+1                                      k       k+1
                          k=0                             k=0

           n−1                       n−1
      2n X n−k n
                  + k+1n   2n+1   2n X                          1        2n+1        2n+1
   =             n−1
                        +      =                               n−1 +
                                                                             = xn +      .
     n + 1 k=0    k
                           n+1    n k=0                         k
                                                                         n+1         n+1
This implies (2) for n + 1.
Problem 4. Let f : [a, b] → [a, b] be a continuous function and let p ∈ [a, b].
Define p0 = p and pn+1 = f (pn ) for n = 0, 1, 2, . . . Suppose that the set
Tp = {pn : n = 0, 1, 2, . . . } is closed, i.e., if x ∈
                                                      / Tp then there is a δ > 0 such
that for all x0 ∈ Tp we have |x0 − x| ≥ δ. Show that Tp has finitely many
elements.
Solution. If for some n > m the equality pm = pn holds then Tp is a finite
set. Thus we can assume that all points p0 , p1 , . . . are distinct. There is
a convergent subsequence pnk and its limit q is in Tp . Since f is continu-
ous pnk +1 = f (pnk ) → f (q), so all, except for finitely many, points pn are
accumulation points of Tp . Hence we may assume that all of them are ac-
cumulation points of Tp . Let d = sup{|pm − pn | : m, n ≥ 0}. Let δn be

                                                2
positive numbers such that ∞            d
                             P
                               n=0 δn < 2 . Let In be an interval of length less
than δn centered at pn such that there are there are infinitely many k’s such
           n
           [
that pk ∈/   Ij , this can be done by induction. Let n0 = 0 and nm+1 be the
          j=0
                                            nm
                                            [
smallest integer k > nm such that pk ∈
                                     /            Ij . Since Tp is closed the limit
                                            j=0
of the subsequence (pnm ) must be in Tp but it is impossible because of the
definition of In ’s, of course if the sequence (pnm ) is not convergent we may
replace it with its convergent subsequence. The proof is finished.
Remark. If Tp = {p1 , p2 , . . . } and each pn is an accumulation point of Tp ,
then Tp is the countable union of nowhere dense sets (i.e. the single-element
sets {pn }). If T is closed then this contradicts the Baire Category Theorem.
Problem 5. Prove or disprove the following statements:
(a) There exists a monotone function f : [0, 1] → [0, 1] such that for each
y ∈ [0, 1] the equation f (x) = y has uncountably many solutions x.
(b) There exists a continuously differentiable function f : [0, 1] → [0, 1] such
that for each y ∈ [0, 1] the equation f (x) = y has uncountably many solutions
x.
Solution. a. It does not exist. For each y the set {x : y = f (x)} is either
empty or consists of 1 point or is an interval. These sets are pairwise disjoint,
so there are at most countably many of the third type.
b. Let f be such a map. Then for each value y of this map there is an x0 such
that y = f (x) and f 0 (x) = 0, because an uncountable set {x : y = f (x)}
contains an accumulation point x0 and clearly f 0 (x0 ) = 0. For every ε > 0
and every x0 such that f 0 (x0 ) = 0 there exists an open interval Ix0 such
that if x ∈ Ix0 then |f 0 (x)| < ε. The union of all these intervals Ix0 may
be written as a union of pairwise disjoint open intervals Jn . The image of
each Jn is an interval (or a point) of length < ε · length(Jn ) due to Lagrange
Mean Value Theorem. Thus the image of the interval [0, 1] may be covered
with the intervals such that the sum of their lengths is ε · 1 = ε. This is not
possible for ε < 1.
Remarks. 1. The proof of part b is essentially the proof of the easy part
of A. Sard’s theorem about measure of the set of critical values of a smooth
map.
2. If only continuity is required, there exists such a function, e.g. the first
co-ordinate of the very well known Peano curve which is a continuous map
from an interval onto a square.



                                       3
                                                                          kM xk2
Problem 6. For an n×n matrix M with real entries let kM k =             sup      ,
                                                                 x∈Rn \{0} kxk2
where k · k2 denotes the Euclidean norm on Rn . Assume that an n × n matrix
                                               1
A with real entries satisfies kAk − Ak−1 k ≤ 2002k for all positive integers k.
               k
Prove that kA k ≤ 2002 for all positive integers k.
Solution.
Lemma 1. Let (an )n≥0 be a sequence of non-negative numbers such that
a2k −a2k+1 ≤ a2k , a2k+1 −a2k+2 ≤ ak ak+1 for any k ≥ 0 and lim sup nan < 1/4.
             √
Then lim sup n an < 1.
Proof. Let cl = supn≥2l (n + 1)an for l ≥ 0. We will show that cl+1 ≤ 4c2l .
Indeed, for any integer n ≥ 2l+1 there exists an integer k ≥ 2l such that
                                                                        c2l
n = 2k or n = 2k + 1. In the first case there is a2k − a2k+1 ≤ a2k ≤ (k+1)  2 ≤

 4c2l       4c2l
2k+1
         − 2k+2
                 , whereas in the second case there is a2k+1 − a2k+2 ≤ ak ak+1 ≤
      c2l          4c2l    4c2l
(k+1)(k+2)
              ≤   2k+2
                        − 2k+3
                                .
                                   4c2l
      Hence a sequence (an − n+1        )n≥2l+1 is non-decreasing and its terms are
                                                                  4c2l
non-positive since it converges to zero. Therefore an ≤ n+1            for n ≥ 2l+1 ,
meaning that c2l+1 ≤ 4c2l . This implies that a sequence ((4cl )2 )l≥0 is non-
                                                                       −l


increasing and therefore bounded from above by some number q ∈ (0, 1) since
                                                                    l
all its terms except finitely many are less than 1. Hence cl ≤ q 2 for l large
                                                             cl       l    √
enough. For any n between 2l and 2l+1 there is an ≤ n+1         ≤ q 2 ≤ ( q)n
                 √       √                        √        √
yielding lim sup n an ≤ q < 1, yielding lim sup n an ≤ q < 1, which ends
the proof.
Lemma 2. Let T be a linear map from Rn into itself. Assume that
lim sup nkT n+1 − T n k < 1/4. Then lim sup kT n+1 − T n k1/n < 1. In particular
T n converges in the operator norm and T is power bounded.
Proof. Put an = kT n+1 − T n k. Observe that
     T k+m+1 − T k+m = (T k+m+2 − T k+m+1 ) − (T k+1 − T k )(T m+1 − T m )
implying that ak+m ≤ ak+m+1 + ak am . Therefore the sequence (am )m≥0 sat-
isfies assumptions of Lemma 1 and the assertion of Proposition 1 follows.
Remarks. 1. The theorem proved above holds in the case of an operator
T which maps a normed space X into itself, X does not have to be finite
dimensional.
2. The constant 1/4 in Lemma 1 cannot be replaced by any greater number
                        1
since a sequence an = 4n  satisfies the inequality ak+m − ak+m+1 ≤ ak am for
any positive integers k and m whereas it does not have exponential decay.
3. The constant 1/4 in Lemma 2 cannot be replaced by any number greater
that 1/e. Consider an operator (T f )(x) = xf (x) on L2 ([0, 1]). One can easily

                                         4
check that lim sup kT n+1 − T n k = 1/e, whereas T n does not converge in the
operator norm. The question whether in general lim sup nkT n+1 − T n k < ∞
implies that T is power bounded remains open.
    Remark The problem was incorrectly stated during the competition: in-
                                            1
stead of the inequality kAk− Ak−1  k ≤ 2002k , the
                                                   inequality
                                                              kAk − Ak−1 k ≤
  1                           1 ε                  1 kε
2002n
      was assumed. If A =             then Ak =           . Therefore
                            0 1                  0 1
                 0 ε
Ak − Ak−1 =            , so for sufficiently small ε the condition is satisfied
                 0 0
although the sequence kAk k is clearly unbounded.
                              




                                      5
