                    IMC 2012, Blagoevgrad, Bulgaria
                                  Day 2, July 29, 2012



Problem 1. Consider a polynomial
                         f (x) = x2012 + a2011 x2011 + . . . + a1 x + a0 .
Albert Einstein and Homer Simpson are playing the following game. In turn, they choose
one of the coefficients a0 , . . . , a2011 and assign a real value to it. Albert has the first move.
Once a value is assigned to a coefficient, it cannot be changed any more. The game ends
after all the coefficients have been assigned values.
    Homer’s goal is to make f (x) divisible by a fixed polynomial m(x) and Albert’s goal
is to prevent this.
 (a) Which of the players has a winning strategy if m(x) = x − 2012?
 (b) Which of the players has a winning strategy if m(x) = x2 + 1?
                                                                                         (10 points)


Problem 2. Define the sequence a0 , a1 , . . . inductively by a0 = 1, a1 = 12 and
                                            na2n
                                an+1 =                   for n ≥ 1.
                                       1 + (n + 1)an
                       ∞
                       X ak+1
Show that the series              converges and determine its value.
                       k=0
                             ak
                                                                                         (10 points)


Problem 3. Is the set of positive integers n such that n! + 1 divides (2012n)! finite or
infinite?
                                                                           (10 points)


Problem 4. Let n ≥ 2 be an integer. Find all real numbers a such that there exist real
numbers x1 , . . . , xn satisfying
              x1 (1 − x2 ) = x2 (1 − x3 ) = . . . = xn−1 (1 − xn ) = xn (1 − x1 ) = a.
                                                                                         (10 points)


Problem 5. Let c ≥ 1 be a real number. Let G be an abelian group and let A ⊂ G be
a finite set satisfying |A + A| ≤ c|A|, where X + Y := {x + y | x ∈ X, y ∈ Y } and |Z|
denotes the cardinality of Z. Prove that
                                                          k
                                   |A   {z. . . + A} | ≤ c |A|
                                    | +A+
                                          k times

for every positive integer k.
                                                                                         (10 points)
