                                             IMC 2024
                                Second Day, August 8, 2024
                                               Solutions


Problem 6. Prove that for any function f : Q → Z, there exist a, b, c ∈ Q such that a < b < c,
f (b) ≥ f (a), and f (b) ≥ f (c).
                 (proposed by Mehdi Golafshan & Markus A. Whiteland, University of Liège, Liège)

Solution 1. We can replace f (x) by the function g(x) = f (1 − x), so without loss of generality we
can assume f (0) ≤ f (1).
   If f (1) ≥ f (2) then we can choose (a, b, c) = (0, 1, 2). Otherwise we have f (0) ≤ f (1) < f (2).
   If there is some x ∈ (1, 2) such that f (x) ≥ f (2) then we can chose (a, b, c) = (1, x, 2); similarly, if
there is some x ∈ (1, 2) with f (x) ≤ f (1) then choose (a, b, c) = (0, 1, x). Hence, in the remaining
cases we have f (1) ≤ f (x) ≤ f (2) for all x ∈ (1, 2).
   Now f is bonded on the interval [1, 2], so it has only finitely many values on this interval. Since
there are infintely many rational numbers in [0, 1], there is a value y that is attained infinitely many
times. The we can choose 1 ≤ a < b < c ≤ 2 such that f (a) = f (b) = f (c) = y.

Solution 2. Assume towards a contradiction that there is a function f which does not satisfy the
claim: for all rationals a, b, c with a < b < c we have f (b) < f (a) or f (b) < f (c).
    Let x and y be arbitrary rationals with x < y. Let I(x, y) = [x, y] ∩ Q. We first observe that
inf f (I(x, y)) = −∞. Indeed, if the infimum was finite, then, as the set f (I(x, y)) is bounded
(sup f (I(x, y)) = max{f (x), f (y)}) and thus finite, there are three points having the same value under
f , which leads to a contradiction regarding our assumption on f .
    So, going back to the question at hand, let x, b, y be arbitrary rationals with x < b < y. Applying
the above observation to the set I(x, b), there exists a point a ∈ I(x, b) such that f (a) < f (b).
Similarly, there exists a point c ∈ I(b, y) such that f (c) < f (b). Hence we have the points a, b, c with
a < b < c and f (b) > max{f (a), f (c)}, which contradicts our assumption on f .




                                                     1
Problem 7. Let n be a positive integer. Suppose that A and B are invertible n × n matrices with
complex entries such that A + B = I (where I is the identity matrix) and

                                              (A2 + B 2 )(A4 + B 4 ) = A5 + B 5 .

Find all possible values of det(AB) for the given n.
                   (proposed by Sergey Bondarev, Sergey Chernov, Belarusian State University, Minsk)

Hint: Find a polynomial p(x) such that p(AB) = 0.

Solution 1. Notice first that AB = A(I − A) = A − A2 = (I − A)A = BA, so A and B commute.
   Let C = AB = BA; then

                  A2 + B 2 = (A + B)2 − 2AB = I − 2C,
                  A4 + B 4 = (A + B)4 − 4AB(A + B)2 + 2A2 B 2 = I − 4C + 2C 2 ,
                  A5 + B 5 = (A + B)5 − 5AB(A + B)3 + 5A2 B 2 (A + B) = I − 5C + 5C 2 ,

so

           0 = (A5 + B 5 ) − (A2 + B 2 )(A4 + B 4 ) = (I − 5C + 5C 2 ) − (I − 2C)(I − 4C + 2C 2 )
             = 4C 3 − 5C 2 + C = 4C(C − I)(C − 41 I);

since C is invertible, we have
                                                        (C − I)(C − 41 I) = 0.
Hence, the polynomial p(x) = (x − 1)(x − 14 ) annihilates the matrix C = AB and therefore all
eigenvalues of C are roots of p(x), so the possible eigenvalues are 1 and 14 . The determinant is the
product of the n eigenvalues, so
                                                    n                   o
                                                        1 1           1
                               det(AB) = det C ∈ 1, 4 , 42 , . . . , 4n .

     Now show that these values are indeed possible.
     If
                                                                                                                
         A = diag 21 , . . . , 12 , eiπ/3 , . . . , eiπ/3   and B = diag 12 , . . . , 12 , e−iπ/3 , . . . , e−iπ/3 ,
                   | {z } |                  {z         }                 | {z } |                   {z          }
                             k                n−k                                             k     n−k

                                                                       
                                        1
then A + B = I, AB = diag               4
                                          , . . . , 14 , 1, . . . , 1       and det(AB) = 41k .
                                        | {z }           | {z }
                                             k             n−k




                                                                            2
 Problem 8. Define the sequence x1 , x2 , . . . by the initial terms x1 = 2, x2 = 4, and the recurrence
 relation
                                                         2n
                              xn+2 = 3xn+1 − 2xn +             for n ≥ 1.
                                                         xn
                xn
 Prove that lim n exists and satisfies
            n→∞ 2
                                            √
                                       1+ 3              xn     3
                                                 ≤ lim n ≤ .
                                          2         n→∞  2      2

(proposed by Karen Keryan, Yerevan State University & American University of Armenia, Armenia)

 Hint: Prove that 2xn ≤ xn+1 ≤ 2xn + n.

 Solution. Let’s prove by induction that xn+1 ≥ 2xn . It holds for n = 1. Assume it holds for n.
 Then by the induction hypothesis we have that xn ≥ 2xn−1 ≥ . . . ≥ 2n−1 x1 > 0 and
                                                               2n
                             xn+2 = 2xn+1 + (xn+1 − 2xn ) +       > 2xn+1 .
                                                               xn
 Similarly we prove that xn+1 ≤ 2xn + n. Again it holds for n = 1. Assume that the inequality holds
 for n. Then using that xn ≥ 2n and the induction hypothesis we obtain

              xn+2 ≤ 3xn+1 − 2xn + 1 ≤ 2xn+1 + (2xn + n) − 2xn + 1 = 2xn+1 + n + 1.
                                                                    xn
    Using the previous inequalities we obtain that the sequence yn = n is increasing and yn+1 ≤
                                                                    2
                                                    xn
 yn + 2nn ≤ . . . ≤ y1 + nk=1 2kk < ∞, thus lim yn = n = c exists.
                        P
                                            n→∞     2
    The recurrence relation has the following form for yn :
                                                                   1
                                4yn+2 − 2yn+1 = 4yn+1 − 2yn +              .
                                                                 2n · yn
 By summing up the above equality for n = 1, . . . , m we obtain
                                                       m          m
                                                       X 1       X      1
                      4ym+2 − 2ym+1 = 4y2 − 2y1 +      n
                                                             =2+      n
                                                                            .                             (1)
                                                  n=1
                                                      2 · yn     n=1
                                                                     2 · yn

 Now using the facts that y1 = 1, yn increases and limn→∞ yn = c we obtain 1 ≤ yn ≤ c. Hence
                                               ∞
                                          1 X 1
                                            ≤      n·y
                                                        ≤ 1.
                                          c   n=1
                                                  2   n


 Thus we get from (1)
                                                           ∞             
                                                           X 1         1
                      2c = lim (4ym+2 − 2ym+1 ) = 2 +             ∈ 2 + ,3 .
                          m→∞
                                                      n=1
                                                          2n · yn      c
                                                                               √
 So we have 2c2 ≥ 2c + 1 and 2c ≤ 3. Recall that c ≥ 1. Therefore 1 +              3 ≤ 2c ≤ 3, which finishes
 the proof.




                                                   3
Problem 9. A matrix A = (aij ) is called nice, if it has the following properties:
 (i) the set of all entries of A is {1, 2, . . . , 2t} for some integer t;
 (ii) the entries are non-decreasing in every row and in every column: ai,j ≤ ai,j+1 and ai,j ≤ ai+1,j ;
(iii) equal entries can appear only in the same row or the same column: if ai,j = ak,ℓ , then either
      i = k or j = ℓ;
 (iv) for each s = 1, 2, . . . , 2t − 1, there exist i ̸= k and j ̸= ℓ such that ai,j = s and ak,ℓ = s + 1.
Prove that for any positive integers m and n, the number          ofnice m× n matrices
                                                                                           is even.
                                                             1 1 1           1 1 3
For example, the only two nice 2 × 3 matrices are                      and             .
                                                             2 2 2           2 4 4
                                            (proposed by Fedor Petrov, St Petersburg State University)

Solution. Define a standard Young tableaux of shape m × n as an m × n matrix with the set of
entries {1, 2, . . . , mn}, increasing in every row and in every column as in (ii).
    Call two standard Young tableaux Y1 , Y2 friends, if they differ by a switch of two consecutive
numbers x, x + 1 (the places of x and x + 1 must be not neighbouring, for such a switch preserving
the monotonicity in rows and columns).
    For a nice m × n matrix A we construct a standard Young tableaux YA of shape m × n as follows:
if A has ni entries equal to i (i = 1, 2, . . . , 2t), we replace them by the numbers from n1 + . . . + ni−1 + 1
to n1 + . . . + ni preserving monotonicity.
    Note that our YA has exactly 2t − 1 friends, where 2t is the number of distinct entries in A, and
moreover, every standard Young tableaux with odd number of friends corresponds to a unique nice
matrix. It remains to apply the handshaking lemma (i.e., the sum of the degrees equals twice the
number of edges in this graph).




                                                       4
Problem 10. We say that a square-free positive integer n is almost prime if
                                       n | xd1 + xd2 + ... + xdk − kx
for all integers x, where 1 = d1 < d2 < ... < dk = n are all the positive divisors of n. Suppose that r
                                                    m
is a Fermat prime (i.e. it is a prime of the form 22 + 1 for an integer m ≥ 0), p is a prime divisor of
an almost prime integer n, and p ≡ 1 (mod r). Show that, with the above notation, di ≡ 1 (mod r)
for all 1 ≤ i ≤ k.
(An integer n is called square-free if it is not divisible by d2 for any integer d > 1.)
                      (proposed by Tigran Hakobyan, Yerevan State University, Vanadzor, Armenia)

Solution. We first prove the following claims.
Lemma 1. If n is almost prime then gcd(n, φ(n)) = 1.
Proof. Assume to the contrary that gcd(n, φ(n)) > 1 so that there are primes p and q dividing n such
that p ≡ 1(mod q). For 0 ≤ i ≤ p − 2 let hi be the number of positive divisors of n congruent to i
modulo p−1 and similarly for 0 ≤ j ≤ q −1 let νj denote the number of positive divisors of n congruent
to j modulo q. Observe that the polynomial Fn (x) = xd1 + xd2 + ... + xdk − kx definesPthe zero function
on Fp due to the condition of the problem. On the other hand, Fn (x) = (h1 − k)x + i̸=1 hi xi in Fp [x],
so that p|hi for all 0 ≤ i ≤ p − 2, i ̸= 1. It follows that 2ω(n)−1 = ν0 = h0 + hq + h2q + ... ≡ 0(mod p)
which is a contradiction (here ω(n) means the number of distinct prime divisors of n). Therefore our
assumption was wrong and the lemma is proved. 2
Lemma 2. Let q be a prime number and let h be a positive integer coprime to q − 1. If l is the order
                                                        l
of h modulo q − 1, then there exists a ∈ Fq such that ah = a and
                                                 2                          l−1
                                  a − ah + ah − ... + (−1)l−1 ah                   ̸= 0
                         l
Proof. Observe that ah = a for any a ∈ Fq since q − 1|hl − 1. On the other hand, the numbers
h0 , h1 , ..., hl−1 leave different remainders upon division by q − 1 and therefore the polynomial
                                                            2                         l−1
                                f (x) = x − xh + xh − ... + (−1)l−1 xh
defines a function on Fq , which is not identically zero. Hence the existence of an element with the
required properties is proved. 2
Lemma 3. If n is almost prime then for any primes p and q dividing n, the order of p modulo q − 1
is an odd number.
Proof. Observe that due to Lemma 1 the order l of p modulo q − 1 is well defined and assume to the
                                                                                             l
contrary that l is an even number. According to Lemma 2 there exists a ∈ Fq such that ap = a and
                                       2                l−1
f (a) ̸= 0, where f (x) = x − xp + xp − ... + (−1)l−1 xp . Let us consider the sequence (ai )li=0 ⊂ Fq
defined by a0 = a and ai+1 = −api for 0 ≤ i ≤ l − 1. Notice that since l is even by the assumption, we
             l
have al = ap0 = a0 . It follows that
                                                     
                   Xl−1 X         l−1
                                  X     X       X pd        l−1 X 
                                                            X                   
                            adi =        adi +   ai  =            adi+1 + apd
                                                                             i    = 0,
                   i=0 d|n       i=0    d| n
                                           p
                                                     d| n
                                                        p
                                                                        i=0 d| n
                                                                               p


since d is always odd being a divisor of n (Recall that gcd(n, φ(n)) = 1 due to Lemma 1, so that
                                                        p
                                           P ai+1d = −ai for all 0 ≤ i ≤ l − 1. On the other hand,
n is odd, except the trivial case n = 2), and
according to the condition of the problem, d|n ai = kai in Fq for all i, which shows that
                                           l−1
                                           X            l−1
                                                        X               l−1 X
                                                                        X
                              kf (a) = k         ai =           kai =              adi = 0
                                           i=0          i=0             i=0 d|n


                                                            5
in Fq which is impossible, since f (a) ̸= 0 by construction and k = 2ω(n)−1 is coprime to q. The
attained contradiction shows that our assumption was wrong and concludes the proof of the lemma.
2
                                                                       m
    Let us get back to the problem. Suppose that p|n is prime and r = 22 + 1 is a Fermat’s prime such
that p ≡ 1(mod r). If q is any prime divisor of n, then by Lemma 3 we have that q l ≡ 1(mod p − 1)
for some odd l, so that q l ≡ 1(mod r) and therefore q = q gcd(l,r−1) ≡ 1(mod r). Hence d ≡ 1(mod r)
for any divisor d of n. 2




                                                 6
