                       IMC 2017, Blagoevgrad, Bulgaria


                                   Day 2, August 3, 2017




Problem 6.    Let f : [0; +∞) → R be a continuous function such that lim f (x) = L exists
                                                                    x→+∞
(it may be nite or innite). Prove that
                                                 Z1
                                           lim        f (nx) dx = L.
                                        n→∞
                                                 0

                                                                                           (10 points)

Problem 7.       Let p(x) be a nonconstant polynomial with real coecients. For every positive
integer n, let
                                 qn (x) = (x + 1)n p(x) + xn p(x + 1).
   Prove that there are only nitely many numbers n such that all roots of qn (x) are real.
                                                                                   (10 points)

Problem 8.       Dene the sequence A1 , A2 , . . . of matrices by the following recurrence:
                                                               
                             0 1                         An I2n
                       A1 =        ,        An+1 =                     (n = 1, 2, . . .)
                             1 0                         I2n An
where Im is the m × m identity matrix.
    Prove
        thatnAn has n + 1 distinct integer eigenvalues λ0 < λ1 < . . . < λn with multiplicities
 n
 0
    , 1 , . . . , n , respectively.
      n

                                                                                    (10 points)

Problem 9.    Dene the sequence f1 , f2 , . . . : [0, 1) → R of continuously dierentiable functions
by the following recurrence:
                   f1 = 1;         0
                                  fn+1 = fn fn+1         on (0, 1),    and fn+1 (0) = 1.
   Show that lim fn (x) exists for every x ∈ [0, 1) and determine the limit function.
             n→∞
                                                                                    (10 points)

Problem 10.       Let K be an equilateral triangle in the plane. Prove that for every p > 0
there exists an ε > 0 with the following property: If n is a positive integer, and T1 , . . . , Tn are
non-overlapping triangles inside K such that each of them is homothetic to K with a negative
ratio, and
                                    n
                                           area(T` ) > area(K) − ε,
                                    X

                                     `=1
then                                   n
                                              perimeter(T` ) > p.
                                       X

                                        `=1
                                                                                           (10 points)
