                      IMC 2018, Blagoevgrad, Bulgaria


                                      Day 2, July 25, 2018




Problem 6. Let k be a positive integer. Find the smallest positive integer n for which there
exist k nonzero vectors v1 , . . . , vk in Rn such that for every pair i, j of indices with |i − j| > 1
the vectors vi and vj are orthogonal.
      (Proposed by Alexey Balitskiy, Moscow Institute of Physics and Technology and M.I.T.)

Solution.  First we prove that if 2n + 1 ≤ k then no sequence v1 , . . . , vk of vectors can satisfy
the condition. Suppose to the contrary that v1 , . . . , vk are vectors with the required property
and consider the vectors
                                          v1 , v3 , v5 , . . . , v2n+1 .
By the condition these n + 1 vectors should be pairwise orthogonal, but this is not possible
in Rn .
   Next we show a possible construction for every pair k, n of positive integers with 2n ≥ k .
Take an orthogonal basis (e1 , . . . , en ) of Rn and consider the vectors

                     v1 = v2 = e1 ,     v3 = v4 = e2 ,        ...,         v2n−1 = v2n = en .

For every pair (i, j) of indices with 1 ≤ i, j ≤ 2n and |i−j| > 1 the vectors vi and vj are distinct
basis vectors, so they are orthogonal. Evidently the subsequence v1 , v2 , . . . , vk also satises the
same property.
   Hence, such a sequence
                      of vectors exists if and only if 2n ≥ k ; that is, for a xed k , the
                         k
smallest suitable n is     .
                         2

Problem 7.    Let (an )∞
                       n=0 be a sequence of real numbers such that a0 = 0 and

                                a3n+1 = a2n − 8 for n = 0, 1, 2, . . .

Prove that the following series is convergent:
                                            ∞
                                            X
                                                   |an+1 − an |.                                    (1)
                                             n=0

                               (Proposed by Orif Ibrogimov, National University of Uzbekistan)

Solution.   We will estimate the ratio between the terms |an+2 − an+1 | and |an+1 − an |.
   Before doing that, we localize the numbers an ; we prove that
                                                 √
                                      −2 ≤ an ≤ − 4 for n ≥ 1.
                                                 3
                                                                                                    (2)
                                                                                                √
The lower bound simply follows from the recurrence: an = 3 a2n−1 − 8 ≥ √3 −8 = −2. The
                                                                               p

proof of the upper bound can be done   by induction:
                                                √     we have√a1 = −2 < − 3 4, and whenever
−2 ≤ an < 0, it follows that an+1 = 3 a2n − 8 ≤ 3 22 − 8 = − 3 4.
                                    p




                                                        1
  Now compare |an+2 − an+1 | with |an+1 − an |. By applying x3 − y 3 = (x − y)(x2 + xy + y 2 ),
x − y 2 = (x − y)(x + y) and the recurrence,
 2


             (a2n+2 + an+2 an+1 + a2n+1 ) · |an+2 − an+1 | =
                                         = |a3n+2 − a3n+1 | = (a2n+1 − 8) − (a2n − 8) =
                                                            = |an+1 + an | · |an+1 − an |.

On the left-hand side we have
                                a2n+2 + an+2 an+1 + a2n+1 ≥ 3 · 42/3 ;

on the right-hand side
                                          |an+1 + an | ≤ 4.
Hence,                                                          √
                                                                3
                                         4                        4
                      |an+2 − an+1 | ≤     2/3
                                               |a n+1 − a n | =     |an+1 − an |.
                                       3·4                      3
By a trivial induction it follows that
                                                    √
                                                    3
                                                        !n−1
                                                      4
                                |an+1 − an | <               |a2 − a1 |.
                                                    3
                                                                                             √
                                                                                             3
                                                                                                 4
Hence the series n=0 |an+1 −an | can be majorized by a geometric series with quotient   < 1;
                 P∞
                                                                                      3
that proves that the series converges.
Problem 8.    Let Ω = {(x, y, z) ∈ Z3 : y + 1 ≥ x ≥ y ≥ z ≥ 0}. A frog moves along the points
of Ω by jumps of length 1. For every positive integer n, determine the number of paths the frog
can take to reach (n, n, n) starting from (0, 0, 0) in exactly 3n jumps.
            (Proposed by Fedor Petrov and Anatoly Vershik, St. Petersburg State University)
Solution.    Let Ψ = {(u, v) ∈ Z3 : v ≥ 0, u ≥ 2v}. Notice that the map π : Ω → Ψ,
π(x, y, z) = (x + y, z) is a bijection between the two sets; moreover π projects all allowed paths
of the frogs to paths inside the set Ψ, using only unit jump vectors. Hence, we are interested
in the number of paths from π(0, 0, 0) = (0, 0) to π(n, n, n) = (2n, n) in the set Ψ, using only
jumps (1, 0) and (0, 1).
    For every lattice point (u, v) ∈ Ψ, let f (u, v) be the number of paths from (0, 0) to (u, v)
in Ψ with u + v jumps. Evidently we have f (0, 0) = 1. Extend this denition to the points
with v = −1 and 2v = u + 1 by setting
                                 f (u, −1) = 0,     f (2v − 1, v) = 0.                               (3)

To any point (u, v) of Ψ other than the origin, the path can come either from (u − 1, v) or from
(u, v − 1), so

                 f (u, v) = f (u − 1, v) + f (u, v − 1) for (u, v) ∈ Ψ \ {(0, 0)}.                   (4)

If we ignore the boundary condition (3), there is a wide family of functions that satisfy (4);
namely, for every integer c, (u, v) 7→ v+c is such a function, with dening this binomial
                                          u+v

coecient to be 0 if v + c is negative or greater than u + v . 
    Along the line 2v = u + 1 we have u+v         3v−1
                                                        = 2 3v−1  = 2 u+v   . Hence, the function
                                                                         
                                          v
                                              =     v        v−1      v−1
                                                             
                                  ∗               u+v      u+v
                                 f (u, v) =             −2
                                                   v        v−1

                                                    2
satises (3), (4) and f (0, 0) = 1. These properties uniquely dene the function f , so f = f ∗ .
    In particular, the number of paths of the frog from (0, 0, 0) to (n, n, n) is
                                                                 
                                                                  3n
                                                         
                                                  3n     3n        n
                    f (π(n, n, n)) = f (2n, n) =     −2       =        .
                                                   n    n−1     2n + 1

               There exist direct proofs for the formula 3n   /(2n + 1). For instance, we can
                                                                
    Remark.
                                                           n
replicate the well-known proof of the formula for the Catalan numbers using the Cycle Lemma
of Dvoretzky and Motzkin (related to the petrol station replenishment problem). See https:
//en.wikipedia.org/wiki/Catalan_number#Sixth_proof

Problem 9.     Determine all pairs P (x), Q(x) of complex polynomials with leading coecient 1
 such that P (x) divides Q(x)2 + 1 and Q(x) divides P (x)2 + 1.
(Proposed by Rodrigo Angelo, Princeton University and Matheus Secco, PUC, Rio de Janeiro)

Solution. The answer is all pairs (1, 1) and (P, P + i), (P, P − i), where P is a non-constant
monic polynomial in C[x] and i is the imaginary unit.
   Notice that if P |Q2 + 1 and Q|P 2 + 1 then P and Q are coprime and the condition is
equivalent with P Q|P 2 + Q2 + 1.
    Lemma.   If P, Q ∈ C[x] are monic polynomials such that P 2 + Q2 + 1 is divisible by P Q,
then deg P = deg Q.
    Proof.Assume for the sake of contradiction that there is a pair (P, Q) with deg P 6= deg Q.
Among all these pairs, take the one with smallest sum deg P + deg Q and let (P, Q) be such
pair. Without loss of generality, suppose that deg P > deg Q. Let S be the polynomial such
that                                    2    2
                                         P +Q +1
                                                 = S.
                                           PQ
Notice that P a solution of the polynomial equation X 2 − QSX + Q2 + 1 = 0, in variable X .
                                                          Q2 + 1
By Vieta's formulas, the other solution is R = QS − P =           . By R = QS − P , the R is
                                                            P
                                                     Q2 + 1
indeed a polynomial, and because P, Q are monic, R =         is also monic. Therefore the pair
                                                        P
(R, Q) satises the conditions of the Lemma. Notice that deg R = 2 deg Q − deg P < deg P ,
which contradicts the minimality of deg P + deg Q. This contradiction establishes the Lemma.
                                                                             P 2 + Q2 + 1
    By the Lemma, we have that deg(P Q) = deg(P 2 + Q2 + 1) and therefore                 is
                                                                                  PQ
a constant polynomial. If P and Q are constant polynomials, we have P = Q = 1. Assuming
that deg P = deg Q ≥ 1, as P and Q are monic, the leading coecient of P 2 + Q2 + 1 is 2
                                                     P 2 + Q2 + 1
and the leading coecient of P Q is 1, which give us              = 2. Finally we have that
                                                          PQ
P 2 + Q2 + 1 = 2P Q and therefore (P − Q)2 = −1, i.e Q = P + i or Q = P − i. It's easy to
check that these pairs are indeed solutions of the problem.




                                                 3
Problem 10.        For R > 1 let DR = {(a, b) ∈ Z2 : 0 < a2 + b2 < R}. Compute
                                                          X         (−1)a+b
                                                  lim                       .
                                                 R→∞                a2 + b2
                                                        (a,b)∈DR


(Proposed by Rodrigo Angelo, Princeton University and Matheus Secco, PUC, Rio de Janeiro)

Solution.    Dene ER = {(a, b) ∈ Z2 \ {(0, 0)} : a2 + b2 < R and a + b is even}. Then
                              X        (−1)a+b      X    1      X                                 1
                                                = 2           −                                       .                 (5)
                                       a2 + b 2       a2 + b2                              a2 + b 2
                            (a,b)∈DR                    (a,b)∈ER                (a,b)∈DR


But a + b is even if and only if one can write (a, b) = (m − n, m + n), and such m, n are unique.
Notice also that a2 + b2 = (m − n)2 + (m + n)2 = 2m2 + 2n2 , hence a2 + b2 < R if and only if
m2 + n2 < R/2. With that we get:
                 X            1               X                  1                                X          1
            2                     =2                                       =                                        .   (6)
                           a + b2
                            2                           (m − n) + (m + n)2
                                                               2                                          m2 + n2
                (a,b)∈ER                  (m,n)∈DR/2                                      (m,n)∈DR/2


   Replacing (6) in (5), we obtain
                                       X      (−1)a+b                  X              1
                                                       =−                                     ,
                                              a2 + b 2                             a2 + b 2
                                   (a,b)∈DR                        R/2≤a2 +b2 <R


where the second sum is evaluated for a and b integers.
   Denote by N (r)pthe number of √ lattice points in the open disk x + y < r . Along the circle
                                                                    2   2     2

with radius r with R/2 ≤ r < R, there are N (r + 0) − N (r − 0) lattice points; each of them
contribute r12 in the sum (7). So we can re-write the sum as a Stieltjes integral:
                                                                  Z √R
                                           X              1              1
                                                                =  √         dN (r).
                                                       a2 + b 2      R/2 r
                                                                           2
                                       R/2≤a2 +b2 <R


It is well-known that N (r) = πr2 + O(r). (Putting a unit square around each lattice point,
these squares cover the disk with radius r − 1 and lie inside the disk with radius r + 1, so there
their total area is between π(r − 1)2 and π(r + 1)2 ). By integrating by parts,
                 Z √R                         √ R      Z √R
                        1              1                         2
                  √       2
                            dN (r) = 2 N (r) √         + √        3
                                                                    N (r) dr
                    R/2 r              r          R/2       R/2 r
                                      2            √ R       Z √R
                                       πr + O(r)                       πr2 + O(r)
                                   =                 √     + 2  √                 dr
                                           r2          R/2         R/2       r3
                                         Z √R             √                    √ 
                                                dr
                                   = 2π √           + O 1/ R = π log 2 + O 1/ R .
                                            R/2 r


   Therefore,
                                                                            Z √R
            X        (−1)a+b                      X              1                 1
    lim                       = − lim                                = − lim √         dN (r) = −π log 2.
   R→∞               a2 + b 2    R→∞                           2
                                                              a +b 2    R→∞    R/2 r
                                                                                     2
          (a,b)∈DR                            R/2≤a2 +b2 <R




                                                              4
